---
title: "sp_harmo"
author: "Shahar Chaikin"
date: "2025-05-04"
output: html_document
---

Here I will harmonize species names within BioTIME 2.0 and BioShifts.
After that I will inspect how many species appear in both datasets.

libraries
```{r}
library(tidyverse)
library(purrr)
```

#1) Bioshifts
##1.1) Data
```{r}
#Bioshifts
bioshifts_raw=read.csv("C:\\Users\\User\\Desktop\\research\\data\\BioShifts_2/BioShiftsToShare/BIOSHIFTS_v3(10_19_2025).csv")
```

##1.2) taxa harmonization
Harmonization of scientific names by correcting spelling errors and converting nomenclatural synonyms to currently accepted names
```{r}
Harmonization_biosh=bdc::bdc_query_names_taxadb(
  unique(bioshifts_raw$sp_name_checked),
  suggest_names = T,ncores = 4)

#save
#write.csv(Harmonization_biosh,"./stage_1_species_harmo/Harmonization_biosh_v2.csv",row.names = F)

#read
Harmonization_biosh=read.csv("./stage_1_species_harmo/Harmonization_biosh_v2.csv")
#saveRDS(Harmonization_biosh,"./stage_1_species_harmo/Harmonization_biosh_v2.rds")
```

Update files with harmonized species
```{r}
# 1. Select the necessary columns from the harmonization results
harmonization_clean <- Harmonization_biosh %>%
  select(
    original_search,
    scientificName,
    taxonomicStatus,
    notes,
    taxonID, kingdom, phylum, class, order, family 
  )

# 2. Join the datasets while keeping original columns
bioshifts_harmonized <- bioshifts_raw %>%
  left_join(
    harmonization_clean, 
    by = c("sp_name_checked" = "original_search")
  ) %>%
  # 3. Create the harmonized name and merge the taxonomic columns
  mutate(
    # Fallback for the species name
    species_harmonized = case_when(
      !is.na(scientificName) ~ scientificName,
      TRUE ~ str_replace_all(sp_name_checked, "_", " ")
    ),
    
    # Fallback for taxonomy: prioritize harmonized (.y), fallback to original (.x)
    kingdom = coalesce(kingdom.y, kingdom.x),
    phylum  = coalesce(phylum.y, phylum.x),
    class   = coalesce(class.y, class.x),
    family  = coalesce(family.y, family.x),
    taxonID = coalesce(taxonID.y, taxonID.x)
  ) %>%
  # 4. Clean up: remove the messy .x and .y suffix columns
  select(
    -ends_with(".x"), 
    -ends_with(".y")
  )

# Check the results to ensure the NAs were filled correctly
bioshifts_harmonized %>% 
  select(sp_name_checked, species_harmonized, class, family, notes) %>% 
  head(10)

# Save the updated dataset
saveRDS(bioshifts_harmonized, "./stage_1_species_harmo/bioshifts_harmonized.rds")
```

#2) BioTIME
##2.1) Data
```{r}
#BioTIME 2.0
biotime_raw=read_rds("C:\\Users\\User\\Desktop\\research\\data\\BioTIME 2\\biotime_v2_query_15April25.rds")
```

##2.2) taxa harmonization
Harmonization of scientific names by correcting spelling errors and converting nomenclatural synonyms to currently accepted names

```{r}
# 1. Extract and clean unique species names from BioTIME
unique_species_biotime <- unique(biotime_raw$valid_name)

# Remove NA values to prevent errors in the harmonization function
unique_species_biotime <- unique_species_biotime[!is.na(unique_species_biotime)]

# 2. Define chunk size and split the vector
# 5,000 is usually a safe balance between speed and stability
chunk_size <- 5000 
species_chunks <- split(unique_species_biotime, ceiling(seq_along(unique_species_biotime) / chunk_size))

# Create the directory if it doesn't exist to avoid write errors
dir.create("./stage_1_species_harmo", showWarnings = FALSE)

# 3. Process in chunks with intermediate saving
harmonized_list <- list()

for (i in seq_along(species_chunks)) {
  message(sprintf("Processing chunk %d of %d...", i, length(species_chunks)))
  
  # Run the bdc harmonization for the current chunk
  chunk_result <- bdc::bdc_query_names_taxadb(
    species_chunks[[i]],
    suggest_names = TRUE,
    ncores = 4 
  )
  
  # Store in the list
  harmonized_list[[i]] <- chunk_result
  
  # Save an intermediate CSV just in case the process crashes later
  write.csv(
    chunk_result, 
    sprintf("./stage_1_species_harmo/biotime_chunk_%d.csv", i), 
    row.names = FALSE
  )
}

# 4. Combine all processed chunks into a single dataframe
Harmonization_biotime <- bind_rows(harmonized_list)

# 5. Save the final compiled dataset
saveRDS(Harmonization_biotime, "./stage_1_species_harmo/Harmonization_biotime_v2.rds")
message("BioTIME harmonization complete and saved!")
```

Merge the harmonized species names into BIOTIME
```{r}
# 1. Select the necessary columns from the BioTIME harmonization results
harmonization_biotime_clean <- Harmonization_biotime %>%
  select(
    original_search,
    scientificName,
    taxonomicStatus,
    notes,
    taxonID, kingdom, phylum, class, order, family 
  )

# 2. Join the harmonization data to the original BioTIME dataset
biotime_harmonized <- biotime_raw %>%
  left_join(
    harmonization_biotime_clean, 
    by = c("valid_name" = "original_search") # Match raw column to search column
  ) %>%
  # 3. Create a definitive 'species_harmonized' column
  mutate(
    species_harmonized = case_when(
      # Condition A: If a verified scientific name exists, use it
      !is.na(scientificName) ~ scientificName,
      
      # Condition B: If it wasn't found (NA), fall back to the original valid_name.
      # Replacing underscores with spaces just in case, ensuring uniform formatting.
      TRUE ~ str_replace_all(valid_name, "_", " ")
    )
  )

# Check the results to verify the merge was successful
biotime_harmonized %>% 
  select(valid_name, species_harmonized, class, family, notes) %>% 
  head(10)

# Save the finalized, harmonized BioTIME dataset
saveRDS(biotime_harmonized, "./stage_1_species_harmo/biotime_harmonized.rds")
```
