---
title: "sp_harmo"
author: "Shahar Chaikin"
date: "2025-05-04"
output: html_document
---

Here I will harmonize species names within BioTIME 2.0 and BioShifts.
After that I will inspect how many species appear in both datasets.

libraries
```{r}
library(tidyverse)
```

#1) Bioshifts
##1.1) Data
```{r}
#Bioshifts
bioshifts_raw=read.csv("C:\\Users\\User\\Desktop\\research\\data\\Bioshifts\\BioShifts.csv")
```

##1.2) taxa harmonization
Harmonization of scientific names by correcting spelling errors and converting nomenclatural synonyms to currently accepted names
```{r}
# Harmonization_biosh=bdc::bdc_query_names_taxadb(
#   unique(bioshifts_raw$Species),
#   suggest_names = T)

#save
#write.csv(Harmonization_biosh,"Harmonization_biosh.csv",row.names = F)

#read
Harmonization_biosh=read.csv("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts/stage_1_species_harmo/Harmonization_biosh.csv") %>% 
  select(-X)

#Species by status
Harmonization_biosh %>% 
  group_by(notes) %>% 
  summarise(species=n())

#Remove species that have uncertain names
Harmonization_biosh=Harmonization_biosh %>% 
  filter(!notes %in%c("notFound")) %>% 
  drop_na(notes,scientificName)
```

Update files with harmonized species
```{r}
#Add harmonized species and gbif key columns
bioshifts_harmo=bioshifts_raw %>% 
  left_join(Harmonization_biosh %>% 
              select(Species=original_search,
                      species_harmonized=scientificName,
                      usageKey=acceptedNameUsageID),
            by="Species") %>% 
  drop_na(species_harmonized) %>% 
  filter(Gradient%in%"Latitudinal") #keep only latitudinal gradients

#write.csv(bioshifts_harmo,"bioshifts_harmo.csv",row.names = F)
```

#2) BioTIME
##2.1) Data
```{r}
#BioTIME 2.0
biotime_raw=read_rds("C:\\Users\\User\\Desktop\\research\\data\\BioTIME 2\\queryBTv2_April_2024.rds")
```

##2.2) taxa harmonization
Harmonization of scientific names by correcting spelling errors and converting nomenclatural synonyms to currently accepted names
```{r eval=FALSE, include=FALSE}
#chunk one
Harmonization_bioti_1=bdc::bdc_query_names_taxadb(
  unique(biotime_raw$valid_name)[1:10000],
  suggest_names = T)
#chunk two
Harmonization_bioti_2=bdc::bdc_query_names_taxadb(
  unique(biotime_raw$valid_name)[10001:20000],
  suggest_names = T)
#chunk three
Harmonization_bioti_3=bdc::bdc_query_names_taxadb(
  unique(biotime_raw$valid_name)[20001:30000],
  suggest_names = T)
#chunk four
Harmonization_bioti_4=bdc::bdc_query_names_taxadb(
  unique(biotime_raw$valid_name)[30001:40000],
  suggest_names = T)
#chunk five
Harmonization_bioti_5=bdc::bdc_query_names_taxadb(
  unique(biotime_raw$valid_name)[40001:56496],
  suggest_names = T)

#bind all chunks
Harmonization_bioti=bind_rows(Harmonization_bioti_1,
                              Harmonization_bioti_2,
                              Harmonization_bioti_3,
                              Harmonization_bioti_4,
                              Harmonization_bioti_5)

#save
write.csv(Harmonization_bioti,"Harmonization_bioti.csv",row.names = F)
```

Remove species that are nit documented in GBIF
```{r}
#read
Harmonization_bioti=read.csv("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts/stage_1_species_harmo/Harmonization_bioti.csv")
#Species by status
Harmonization_bioti %>% 
  group_by(notes) %>% 
  summarise(species=n())

#Remove species that have uncertain names
Harmonization_bioti=Harmonization_bioti %>% 
  filter(!notes %in%c("notFound")) %>% 
  drop_na(notes,scientificName)
```

Update files with harmonized species
```{r}
#Add harmonized species and gbif key columns
biotime_harmo=biotime_raw %>% 
  left_join(Harmonization_bioti %>% 
              select(valid_name=original_search,
                      species_harmonized=scientificName,
                      usageKey=acceptedNameUsageID),
            by="valid_name") %>% 
  drop_na(species_harmonized)

write_rds(biotime_harmo,"biotime_harmo.rds")
```