---
title: "stage_4_bind_biotime_bioshifts"
author: "Shahar Chaikin"
date: "2026-05-19"
output: html_document
---
# Overview
This script integrates standardized BioTIME abundance time-series with BioShifts range-shift velocities. It uses strict ecological bounding: BioTIME populations are only joined to BioShifts estimates if they physically occur inside the BioShifts study area polygon, and if the majority (>50%) of the BioTIME sampled years fall within the temporal window of the BioShifts range-shift estimation.

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(h3jsr)
library(rnaturalearth)
library(patchwork)
```


#1) Load Data and Shapefiles
Make sure the shapefile path matches where you unzipped the BioShifts polygons.

```{r}
# Load Standardized BioTIME
biotime_ts <- read_rds("./stage_3_std_biotime_intersect/biotime_ts_final.rds")

# Load Harmonized BioShifts
bioshifts_df <- read_rds("./stage_1_species_harmo/bioshifts_harmonized.rds")

# Load BioShifts Polygons (Updated for .gpkg)
# If the gpkg has multiple layers, st_read() will read the first one by default.
# You can specify a layer with st_read("...", layer = "layer_name") if needed.
bioshifts_poly <- st_read("C:/Users/User/Desktop/research/data/BioShifts_2/BioShiftsToShare/ShapefilesBioShiftsv3_studyareas.gpkg") %>%
  # Ensure CRS is WGS84 to match H3 coordinates
  st_transform(4326)
```

#2) Spatial Intersection (Polygons to H3 Cells)
We extract the unique H3 cells from BioTIME, convert them to spatial points, and find which BioShifts polygons they fall into.
```{r}
# Temporarily turn off the strict s2 spherical geometry engine
sf_use_s2(FALSE)

# Clean up any invalid geometries in the BioShifts polygons
bioshifts_poly_clean <- bioshifts_poly %>% 
  st_make_valid()

# Extract unique H3 cells and convert to sf points
h3_points <- biotime_ts %>%
  select(h3_id) %>%
  distinct() %>%
  mutate(geometry = cell_to_point(h3_id)) %>%
  st_as_sf(crs = 4326)

# Spatially join H3 points to the cleaned BioShifts polygons
h3_polygon_map <- st_join(h3_points, bioshifts_poly_clean, join = st_intersects) %>%
  st_drop_geometry() %>%
  # Use 'Name' instead of 'ID' to match the .gpkg structure
  filter(!is.na(Name)) %>%
  # Keep the H3 ID and the Polygon Name
  select(h3_id, Name)

# Turn s2 back on
sf_use_s2(TRUE)

cat("Successfully mapped", nrow(h3_polygon_map), "H3-to-Polygon intersections.\n")
```

# 3) Taxonomic & Temporal Integration
```{r}
# 1. Attach the polygon Name to BioTIME
biotime_mapped <- biotime_ts %>%
  inner_join(h3_polygon_map, by = "h3_id", relationship = "many-to-many")

# 2. Join with BioShifts data (Requires both Spatial match and Species match)
integrated_data <- biotime_mapped %>%
  inner_join(
    bioshifts_df, 
    # Match the shapefile's 'Name' to the CSV's 'ID'
    by = c("Name" = "ID", "species_harmonized"),
    relationship = "many-to-many"
  )

# 3. Apply the >50% Temporal Overlap Rule
final_dataset <- integrated_data %>%
  group_by(unique_pop_id, ID_BioShifts) %>%
  mutate(
    # Total number of sampled years for this population in BioTIME
    total_sampled_years = n(),
    
    # Check if the BioTIME year falls within the BioShifts estimation window
    is_in_window = YEAR >= START_firstperiod & YEAR <= END_secondperiod,
    
    # Count how many years fall in the window
    years_in_window = sum(is_in_window, na.rm = TRUE),
    
    # Calculate percentage
    perc_in_window = years_in_window / total_sampled_years
  ) %>%
  # Keep only populations where strictly more than 50% of sampling occurred during the shift window
  filter(perc_in_window > 0.5) %>%
  ungroup() %>%
  
  # Clean up calculation columns
  select(-is_in_window, -years_in_window)


final_dataset <- final_dataset %>%
  # Drop all the redundant taxonomic columns that came from BioShifts (.y)
  select(-ends_with(".y")) %>%
  # Remove the ".x" suffix from the BioTIME taxonomic columns so they are clean again
  rename_with(~ str_remove(., "\\.x$"), ends_with(".x"))

saveRDS(final_dataset,"./stage_4_bind_biotime_bioshifts/final_dataset.rds")

# Quick Summary
cat("Original BioTIME rows:", nrow(biotime_ts), "\n")
cat("Final Integrated rows:", nrow(final_dataset), "\n")
cat("Unique matched populations:", n_distinct(final_dataset$unique_pop_id), "\n")


# ---------------------------------------------------------
#Extract Target Species List for Occurrence Download
# ---------------------------------------------------------
# We only want to download occurrences for species that made it through the pipeline
target_species_list <- final_dataset %>%
  distinct(species_harmonized, taxonID,kingdom, phylum, class, order, family, Eco) %>%
  arrange(class, species_harmonized)

# Save this as a clean CSV or RDS to load into your new occurrence-fetching script
write_csv(target_species_list, "./stage_4_bind_biotime_bioshifts/target_species_for_occurrences.csv")

cat("Exported", nrow(target_species_list), "unique species for global occurrence extraction.\n")
```

#4) Creating final_dataset_lite
```{r}
# Recreate the lite dataset, keeping ALL required columns for modeling and mapping
final_dataset_lite <- final_dataset %>%
  # 1. Clean the messy .x and .y column names first
  select(-ends_with(".y")) %>%
  rename_with(~ str_remove(., "\\.x$"), ends_with(".x")) %>%
  
  # 2. Select the essential columns (including LonCentDeg this time!)
  select(
    unique_pop_id, h3_id, ID_BioShifts,
    YEAR, mean_annual_abundance, n_samples_available, total_sampled_years,perc_in_window,
    Type, Rate, UnitRate, Eco, Param,STUDY_ID, 
    LatCentDeg, LonCentDeg, # Both coordinates are safely included here
    species_harmonized, kingdom, phylum, class, order, family
  ) %>%
  
  # 3. Create the clean Range_Position column from Param
  mutate(
    Range_Position = case_when(
      Param == "LE" ~ "Leading_Edge",
      Param == "TE" ~ "Trailing_Edge",
      Param == "C"  ~ "Centroid",
      Param == "O"  ~ "Optimum",
      TRUE ~ "Unknown"
    )
  )

#saveRDS(final_dataset_lite,"./stage_4_bind_biotime_bioshifts/final_dataset_lite.rds")
# Check the memory savings
cat("Full dataset size: ", format(object.size(final_dataset), units = "Mb"), "\n")
cat("Lite dataset size: ", format(object.size(final_dataset_lite), units = "Mb"), "\n")

# Preview the clean data
head(final_dataset_lite)

# Check the minimum number of years per population
final_dataset_lite %>%
  group_by(unique_pop_id, ID_BioShifts) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  summary(n_years)
```

# 5. Resolve Overlaps (The "Winning Tickets")
 Clean the dataset by resolving overlapping studies
```{r}

# We create a dictionary of the best BioShifts study per population & shift type
final_dataset_clean <- final_dataset_lite %>%
  
  # 1. Restrict to overall range shift parameters (Centroid or Optimum)
  filter(Param %in% c("C", "O")) %>%
  
  # 2. Group by BOTH the unique BioTIME population AND the shift Type (LAT, ELE, DEP)
  group_by(unique_pop_id, Type) %>%
  
  # 3. Sort by highest temporal overlap percentage
  arrange(desc(perc_in_window)) %>% 
  
  # 4. Keep only the top row (the best matching study)
  slice(1) %>%
  
  ungroup()
```

# 6) Extract the Population Slopes
```{r}
library(broom.mixed)

# 6a. Recover the Full Time-Series for the winning combinations
master_time_series <- final_dataset_lite %>%
  semi_join(final_dataset_clean, by = c("unique_pop_id", "ID_BioShifts", "Type", "Param"))
# Save the clean master dataset
saveRDS(master_time_series, "./stage_4_bind_biotime_bioshifts/master_time_series.rds")
```

Define the Smart Fallback Modeling Function (Tracking Model Type)
```{r}
extract_population_trend <- function(df) {
  
  # Helper function to safely extract coefficients and append the model type
  safe_extract <- function(mod, model_name, is_lm = FALSE) {
    if (is_lm) {
      coefs <- summary(mod)$coefficients
      slope <- coefs["YEAR", "Estimate"]
      se <- coefs["YEAR", "Std. Error"]
    } else {
      coefs <- summary(mod)$coefficients$cond
      slope <- coefs["YEAR", "Estimate"]
      se <- coefs["YEAR", "Std. Error"]
    }
    
    # If the Hessian failed, SE will be NaN. Throw an error to trigger the next fallback.
    if (is.na(se) | is.nan(se)) stop("Invalid Standard Error")
    
    # Return the data with the new model_type column
    return(tibble(slope = slope, se = se, model_type = model_name))
  }
  
  # Check if the time-series actually contains zeros
  has_zeros <- any(df$mean_annual_abundance == 0)
  
  # ATTEMPT 1: Zero-Inflated Negative Binomial (Only run if zeros exist)
  if (has_zeros) {
    res <- tryCatch({
      mod <- glmmTMB(mean_annual_abundance ~ YEAR, ziformula = ~ 1, family = nbinom2(link = "log"), data = df)
      safe_extract(mod, "ZINB")
    }, error = function(e) NULL)
    if (!is.null(res)) return(res)
  }
  
  # ATTEMPT 2: Standard Negative Binomial (Drops zero-inflation overhead)
  res <- tryCatch({
    mod <- glmmTMB(mean_annual_abundance ~ YEAR, family = nbinom2(link = "log"), data = df)
    safe_extract(mod, "NB2")
  }, error = function(e) NULL)
  if (!is.null(res)) return(res)
  
  # ATTEMPT 3: Poisson (Drops the dispersion parameter overhead)
  res <- tryCatch({
    mod <- glmmTMB(mean_annual_abundance ~ YEAR, family = poisson(link = "log"), data = df)
    safe_extract(mod, "Poisson")
  }, error = function(e) NULL)
  if (!is.null(res)) return(res)
  
  # ATTEMPT 4: Gaussian log-linear (The ultimate safety net)
  res <- tryCatch({
    mod <- lm(log1p(mean_annual_abundance) ~ YEAR, data = df)
    safe_extract(mod, "Gaussian_log1p", is_lm = TRUE)
  }, error = function(e) tibble(slope = NA_real_, se = NA_real_, model_type = "Failed"))
  
  return(res)
}

# Run the loop across all populations 
master_slopes_raw <- master_time_series %>%
  group_by(unique_pop_id, Type) %>% 
  nest() %>%
  
  # THE FIX: suppressWarnings goes INSIDE the mutate, wrapping the map function!
  mutate(trend = suppressWarnings(map(data, extract_population_trend))) %>%
  
  unnest(trend) %>%
  select(-data) %>%
  ungroup() %>%
  
  # Drop the extreme edge cases that still failed
  filter(!is.na(slope), !is.na(se)) %>%
  
  # Calculate precision weight (Inverse Variance) for the meta-analysis
  mutate(inv_var_weight = 1 / (se^2))

cat("Rescued dataset now contains slopes for", n_distinct(master_slopes_raw$unique_pop_id), "populations.\n")


# Merge Metadata and Finalize Master Dataset
# Drop time-varying columns from the clean dictionary
metadata_to_merge <- final_dataset_clean %>%
  select(-YEAR, -mean_annual_abundance)

# Join the rich metadata back to the calculated slopes
master_population_slopes <- master_slopes_raw %>%
  left_join(metadata_to_merge, by = c("unique_pop_id", "Type"))

cat("Successfully calculated master slopes for", nrow(master_population_slopes), "populations.\n")

# Save the ultimate master file
saveRDS(master_population_slopes, "./stage_4_bind_biotime_bioshifts/master_population_slopes.rds")
```

#6) Visualizations
```{r}
# ---------------------------------------------------------
# 1. Quick Summary Statistics (The Raw Numbers)
# ---------------------------------------------------------
cat("Total Unique Populations:", n_distinct(final_dataset_clean$unique_pop_id), "\n")
cat("Total Unique Species:", n_distinct(final_dataset_clean$species_harmonized), "\n")
cat("Total BioShifts Polygons Mapped:", n_distinct(final_dataset_clean$ID_BioShifts), "\n")
```

##data 
```{r}
# Calculate the grand totals
total_pops <- n_distinct(final_dataset_clean$unique_pop_id)
total_species <- n_distinct(final_dataset_clean$species_harmonized)

# 1. Prepare data for the taxonomy plot
taxa_data <- final_dataset_clean %>%
  distinct(unique_pop_id, class, Eco) %>%
  filter(!is.na(class), !is.na(Eco)) %>%
  # Count how many populations belong to each class-realm combination
  count(class, Eco) %>%
  # Group by class to calculate the total populations per class
  group_by(class) %>%
  mutate(total_n = sum(n)) %>%
  ungroup() %>%
  # Keep only the top 10 classes based on their total counts
  filter(dense_rank(desc(total_n)) <= 10)

max_pop_n <- max(taxa_data$total_n)

# ---------------------------------------------------------
# PLOT A: Taxonomic Richness
# ---------------------------------------------------------
plot_taxa <- taxa_data %>%
  ggplot(aes(x = reorder(class, total_n), y = n, fill = Eco)) +
  geom_col(color = "black", alpha = 0.8) +
  
  # Text labels for the total count per bar
  geom_text(
    data = taxa_data %>% distinct(class, total_n),
    aes(x = reorder(class, total_n), y = total_n, label = total_n),
    hjust = -0.2, size = 3.5, inherit.aes = FALSE
  ) +
  
  coord_flip() + 
  
  scale_fill_manual(
    values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a"),
    labels = c("Mar" = "Marine", "Ter" = "Terrestrial", "Fre" = "Freshwater")
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+

  theme_bw() +
  labs(
    title = "A) Top 10 classes by population count",
    caption = paste("Unique populations =", total_pops), # Added to caption
    x = "Taxonomic class", 
    y = "Unique populations",
    fill = "Realm"
  ) +
  theme(
    # Move legend inside the plot area (bottom right)
    legend.position = c(0.85, 0.15), 
    # Add a white background with a subtle border so it pops against grid lines
    legend.background = element_rect(fill = "white", color = "grey60", linewidth = 0.3),
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    # Adjust caption formatting to make it look clean
    plot.caption = element_text(size = 10, face = "bold", color = "grey30", hjust = 1)
  ) 

plot_taxa

# ---------------------------------------------------------
# 2. Prepare data for the realm species plot
# ---------------------------------------------------------
eco_species_data <- final_dataset_clean %>%
  distinct(species_harmonized, Eco) %>%
  count(Eco)

max_eco_n <- max(eco_species_data$n)

# ---------------------------------------------------------
# PLOT B: Unique Species by Realm
# ---------------------------------------------------------
plot_eco_species <- eco_species_data %>%
  ggplot(aes(x = Eco, y = n, fill = Eco)) +
  geom_col(color = "black", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f")) +
  scale_x_discrete(labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  theme_bw() +
  labs(
    title = "B) Unique species by realm",
    caption = paste("Unique species =", total_species), # Added to caption
    x = "Realm", 
    y = "Number of unique species"
  ) +
  theme(
    legend.position = "none",
    # Match the caption styling of Plot A
    plot.caption = element_text(size = 10, face = "bold", color = "grey30", hjust = 1)
  )

plot_eco_species

# ---------------------------------------------------------
# 3. Prepare data for the temporal distribution plot
# ---------------------------------------------------------
duration_data <- final_dataset_clean %>%
  distinct(unique_pop_id, total_sampled_years)

median_duration <- median(duration_data$total_sampled_years, na.rm = TRUE)

# ---------------------------------------------------------
# PLOT C: Distribution of Sampled Years
# ---------------------------------------------------------
plot_years <- duration_data %>%
  ggplot(aes(x = total_sampled_years)) +
  # Using a distinct color from your palette to separate it from the realm plots
  geom_histogram(fill = "#e76f51", color = "black", binwidth = 5, alpha = 0.8) +
  
  # Add a dashed line to indicate the median
  geom_linerange(aes(x = median_duration,ymin=0,ymax=2200),
                 linetype = "dashed", color = "grey30", linewidth = 0.8) +
  
  # Annotate the median value near the top of the line
  annotate("text", x = median_duration + 1, y = Inf, 
           label = paste("Median =", median_duration, "years"), 
           hjust = 0, vjust = 2, size = 4, fontface = "bold", color = "grey30") +
  
  theme_bw() +
  labs(
    title = "C) Distribution of sampled years",
    x = "Number of unique years sampled",
    y = "Unique populations"
  ) +
  theme(
    # Keeping styling consistent with the other panels
    plot.title = element_text(hjust = 0)
  )

plot_years
```


Bind
```{r}
library(gridExtra)

combined_panel <- arrangeGrob(
  plot_taxa, 
  arrangeGrob(plot_eco_species, plot_years, ncol = 1), 
  ncol = 2,
  widths = c(1.3, 1) # Gives the taxa plot slightly more horizontal space for the long labels
)

# Display the combined plot in the viewer
grid::grid.draw(combined_panel)

# Save the final 3-panel figure
ggsave("./stage_4_bind_biotime_bioshifts/Taxa_Realm_Duration_Panel.png", 
       plot = combined_panel, 
       width = 12, height = 6, dpi = 300, bg = "white")
```

##Maps
```{r}
library(gridExtra)
library(grid)

## Maps
# 1. Calculate unique species per H3 hexagon
h3_hexagons <- final_dataset_clean %>%
  group_by(h3_id) %>%
  summarise(n_species = n_distinct(species_harmonized), .groups = "drop")

# Extract raw polygons and convert to sf
h3_hexagons$geometry <- cell_to_polygon(h3_hexagons$h3_id, simple = TRUE)
h3_hexagons <- st_as_sf(h3_hexagons, crs = 4326)

# Split polygons that cross the International Date Line
h3_hexagons <- st_wrap_dateline(h3_hexagons)

world_sf <- ne_countries(scale = "large", type = "map_units", returnclass = "sf")

# ---------------------------------------------------------
# 1. Americas Map 
# ---------------------------------------------------------
plot_map_americas <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = n_species), alpha=0.8, color = "black", linewidth = 0.1) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-14500000, -4500000), ylim = c(1000000, 7000000)) +
  scale_fill_viridis_c(
    name = "Unique species", option = "viridis", direction = -1,      
    guide = guide_colorbar(
      direction = "horizontal", barwidth = unit(4, "cm"), barheight = unit(0.5, "cm"),
      title.position = "top", title.hjust = 0.5, ticks.colour = "black", frame.colour = "black" 
    )
  ) +
  theme_minimal() +
  labs(title = "A) North America & Pacific", x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    plot.title.position = "plot",
    # Margin fixed: no more negative margins needed with gridExtra!
    plot.title = element_text(hjust = 0, face = "bold", size = 14, margin = margin(b = 5)),
    axis.text = element_text(size = 8, color = "grey30"),
    legend.position = c(0.2, 0.22), 
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.title = element_text(size = 11, face = "bold"), 
    legend.text = element_text(size = 10),           
    legend.margin = margin(t = 8, r = 10, b = 8, l = 10), 
    plot.margin = margin(10, 10, 10, 10)
  )

# ---------------------------------------------------------
# 2. South Africa Inset (Converted to a grob for embedding)
# ---------------------------------------------------------
plot_map_sa <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = n_species), alpha=0.8, color = "black", 
          linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = 4326, xlim = c(15, 35), ylim = c(-36, -22)) +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  theme_minimal() +
  labs(title = "C) South Africa", x = NULL, y = NULL) +
  scale_y_continuous(breaks = seq(-40,-20, by = 4)) +
  scale_x_continuous(breaks = seq(15,35, by = 10)) +
  theme(
    # Solid white background so Europe doesn't bleed through
    plot.background = element_rect(fill = "white", color = "black", linewidth = 0.5), 
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", size = 10, margin = margin(t = 2, b = 2, l = 2)),
    axis.text = element_text(size = 6, color = "grey30"),
    plot.margin = margin(2, 2, 2, 2)
  )

# Convert the SA map into a graphical object (grob)
sa_grob <- ggplotGrob(plot_map_sa)

# ---------------------------------------------------------
# 3. Europe Focus (With Embedded SA Inset)
# ---------------------------------------------------------
plot_map_europe <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = n_species), alpha=0.8, color = "black", 
          linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-1000000, 2000000), ylim = c(3500000, 7500000)) +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  
  # THE FIX: Embed the SA grob directly into the Europe map coordinates (Bottom Right)
  annotation_custom(
    grob = sa_grob, 
    xmin = 500000, xmax = 2050000,  # Longitude coordinates for the Mediterranean
    ymin = 3300000, ymax = 4700000  # Latitude coordinates for the Mediterranean
  ) +
  
  theme_minimal() +
  labs(title = "B) Europe", x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", size = 14, margin = margin(b = 5)),
    axis.text = element_text(size = 8, color = "grey30"),
    plot.margin = margin(10, 10, 10, 10)
  )
plot_map_europe
# ---------------------------------------------------------
# Assemble and Save via GridExtra
# ---------------------------------------------------------
# arrangeGrob prevents the plot from immediately drawing in the console, saving it straight to the object
final_maps_only <- arrangeGrob(
  plot_map_americas, plot_map_europe, 
  ncol = 2, 
  widths = c(1.6, 1) # Gives the Americas slightly more width to balance the aspect ratios perfectly
)

ggsave("./stage_4_bind_biotime_bioshifts/maps.png", 
       plot = final_maps_only, 
       width = 14, height = 7, dpi = 300, bg = "white")
```