---
title: "stage_4_bind_biotime_bioshifts"
author: "Shahar Chaikin"
date: "2026-05-19"
output: html_document
---
# Overview
This script integrates standardized BioTIME abundance time-series with BioShifts range-shift velocities. It uses strict ecological bounding: BioTIME populations are only joined to BioShifts estimates if they physically occur inside the BioShifts study area polygon, and if the majority (>50%) of the BioTIME sampled years fall within the temporal window of the BioShifts range-shift estimation.

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(h3jsr)
library(rnaturalearth)
library(patchwork)
```


#1) Load Data and Shapefiles
Make sure the shapefile path matches where you unzipped the BioShifts polygons.

```{r}
# Load Standardized BioTIME
biotime_ts <- read_rds("./stage_3_std_biotime_intersect/biotime_ts_final.rds")

# Load Harmonized BioShifts
bioshifts_df <- read_rds("./stage_1_species_harmo/bioshifts_harmonized.rds")

# Load BioShifts Polygons (Updated for .gpkg)
# If the gpkg has multiple layers, st_read() will read the first one by default.
# You can specify a layer with st_read("...", layer = "layer_name") if needed.
bioshifts_poly <- st_read("C:/Users/User/Desktop/research/data/BioShifts_2/BioShiftsToShare/ShapefilesBioShiftsv3_studyareas.gpkg") %>%
  # Ensure CRS is WGS84 to match H3 coordinates
  st_transform(4326)
```

#2) Spatial Intersection (Polygons to H3 Cells)
We extract the unique H3 cells from BioTIME, convert them to spatial points, and find which BioShifts polygons they fall into.
```{r}
# Temporarily turn off the strict s2 spherical geometry engine
sf_use_s2(FALSE)

# Extract unique H3 cells and convert to sf points
h3_points <- biotime_ts %>%
  select(h3_id) %>%
  distinct() %>%
  mutate(geometry = cell_to_point(h3_id)) %>%
  st_as_sf(crs = 4326)

# Spatially join H3 points to the cleaned BioShifts polygons
h3_polygon_map <- st_join(h3_points, bioshifts_poly_clean, join = st_intersects) %>%
  st_drop_geometry() %>%
  # Use 'Name' instead of 'ID' to match the .gpkg structure
  filter(!is.na(Name)) %>%
  # Keep the H3 ID and the Polygon Name
  select(h3_id, Name)

# Turn s2 back on
sf_use_s2(TRUE)

cat("Successfully mapped", nrow(h3_polygon_map), "H3-to-Polygon intersections.\n")
```

# 3) Taxonomic & Temporal Integration
```{r}
# 1. Attach the polygon Name to BioTIME
biotime_mapped <- biotime_ts %>%
  inner_join(h3_polygon_map, by = "h3_id", relationship = "many-to-many")

# 2. Join with BioShifts data (Requires both Spatial match and Species match)
integrated_data <- biotime_mapped %>%
  inner_join(
    bioshifts_df, 
    # Match the shapefile's 'Name' to the CSV's 'ID'
    by = c("Name" = "ID", "species_harmonized"),
    relationship = "many-to-many"
  )

# 3. Apply the >50% Temporal Overlap Rule
final_dataset <- integrated_data %>%
  group_by(unique_pop_id, ID_BioShifts) %>%
  mutate(
    # Total number of sampled years for this population in BioTIME
    total_sampled_years = n(),
    
    # Check if the BioTIME year falls within the BioShifts estimation window
    is_in_window = YEAR >= START_firstperiod & YEAR <= END_secondperiod,
    
    # Count how many years fall in the window
    years_in_window = sum(is_in_window, na.rm = TRUE),
    
    # Calculate percentage
    perc_in_window = years_in_window / total_sampled_years
  ) %>%
  # Keep only populations where strictly more than 50% of sampling occurred during the shift window
  filter(perc_in_window > 0.5) %>%
  ungroup() %>%
  
  # Clean up calculation columns
  select(-is_in_window, -years_in_window, -perc_in_window)


final_dataset <- final_dataset %>%
  # Drop all the redundant taxonomic columns that came from BioShifts (.y)
  select(-ends_with(".y")) %>%
  # Remove the ".x" suffix from the BioTIME taxonomic columns so they are clean again
  rename_with(~ str_remove(., "\\.x$"), ends_with(".x"))

saveRDS(final_dataset,"./stage_4_bind_biotime_bioshifts/final_dataset.rds")

# Quick Summary
cat("Original BioTIME rows:", nrow(biotime_ts), "\n")
cat("Final Integrated rows:", nrow(final_dataset), "\n")
cat("Unique matched populations:", n_distinct(final_dataset$unique_pop_id), "\n")
```

#4) Creating final_dataset_lite
```{r}
# Recreate the lite dataset, keeping ALL required columns for modeling and mapping
final_dataset_lite <- final_dataset %>%
  # 1. Clean the messy .x and .y column names first
  select(-ends_with(".y")) %>%
  rename_with(~ str_remove(., "\\.x$"), ends_with(".x")) %>%
  
  # 2. Select the essential columns (including LonCentDeg this time!)
  select(
    unique_pop_id, h3_id, ID_BioShifts,
    YEAR, mean_annual_abundance, n_samples_available, total_sampled_years,
    Type, Rate, UnitRate, Eco, Param, 
    LatCentDeg, LonCentDeg, # Both coordinates are safely included here
    species_harmonized, kingdom, phylum, class, order, family
  ) %>%
  
  # 3. Create the clean Range_Position column from Param
  mutate(
    Range_Position = case_when(
      Param == "LE" ~ "Leading_Edge",
      Param == "TE" ~ "Trailing_Edge",
      Param == "C"  ~ "Centroid",
      Param == "O"  ~ "Optimum",
      TRUE ~ "Unknown"
    )
  )

saveRDS(final_dataset_lite,"./stage_4_bind_biotime_bioshifts/final_dataset_lite.rds")
# Check the memory savings
cat("Full dataset size: ", format(object.size(final_dataset), units = "Mb"), "\n")
cat("Lite dataset size: ", format(object.size(final_dataset_lite), units = "Mb"), "\n")

# Preview the clean data
head(final_dataset_lite)

# Check the minimum number of years per population
final_dataset_lite %>%
  group_by(unique_pop_id, ID_BioShifts) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  summary(n_years)
```

#5) Visualizations
```{r}
# ---------------------------------------------------------
# 1. Quick Summary Statistics (The Raw Numbers)
# ---------------------------------------------------------
cat("Total Unique Populations:", n_distinct(final_dataset_lite$unique_pop_id), "\n")
cat("Total Unique Species:", n_distinct(final_dataset_lite$species_harmonized), "\n")
cat("Total BioShifts Polygons Mapped:", n_distinct(final_dataset_lite$ID_BioShifts), "\n")
```

##data 
```{r}
# Calculate the grand totals
total_pops <- n_distinct(final_dataset_lite$unique_pop_id)
total_species <- n_distinct(final_dataset_lite$species_harmonized)

# 1. Prepare data for the taxonomy plot to find the max value for text placement
taxa_data <- final_dataset_lite %>%
  distinct(unique_pop_id, class) %>%
  filter(!is.na(class)) %>%
  count(class, sort = TRUE) %>%
  slice_max(n, n = 10)

max_pop_n <- max(taxa_data$n)

# PLOT B: Taxonomic Richness (Populations per Class)
plot_taxa <- taxa_data %>%
  ggplot(aes(x = reorder(class, n), y = n)) +
  geom_col(fill = "#e76f51", color = "black", alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() + 
  # Expand the y-axis a bit more to fit the annotation
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  # ADD TEXT INSIDE PLOT: Positioned in the bottom right corner
  annotate("text", x = 1.5, y = max_pop_n * 1.1, 
           label = paste("Unique populations =", total_pops), 
           hjust = 1, size = 4, fontface = "bold", color = "grey30") +
  theme_minimal() +
  labs(title = "B) Top 10 classes by population count",
       x = "Taxonomic class", y = "Number of populations")

# 2. Prepare data for the ecosystem species plot to find the max value
eco_species_data <- final_dataset_lite %>%
  distinct(species_harmonized, Eco) %>%
  count(Eco)

max_eco_n <- max(eco_species_data$n)

# PLOT D: Unique Species by Ecosystem
plot_eco_species <- eco_species_data %>%
  ggplot(aes(x = Eco, y = n, fill = Eco)) +
  geom_col(color = "black", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  # Expand the y-axis to fit the annotation above the bars
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  # ADD TEXT INSIDE PLOT: Positioned top center
  annotate("text", x = 0.8, y = max_eco_n * 1.1, 
           label = paste("Unique species =", total_species), 
           size = 4, fontface = "bold", color = "grey30") +
  theme_minimal() +
  labs(title = "D) Unique species by ecosystem",
       x = "Ecosystem", y = "Number of unique species") +
  theme(legend.position = "none")
plot_eco_species
```

```{r}
# PLOT D: Unique Species by Ecosystem (CORRECTED)
plot_eco_species <- eco_species_data %>%
  ggplot(aes(x = Eco, y = n, fill = Eco)) +
  # ADDED show.legend = FALSE here
  geom_col(color = "black", alpha = 0.8, show.legend = FALSE) + 
  geom_text(aes(label = n), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  annotate("text", x = 0.8, y = max_eco_n * 1.1, 
           label = paste("Unique species =", total_species), 
           size = 4, fontface = "bold", color = "grey30") +
  theme_minimal() +
  labs(title = "D) Unique species by ecosystem",
       x = "Ecosystem", y = "Number of unique species")
  # Removed theme(legend.position = "none") as it is no longer needed

# PLOT C: Populations by Range Position (CORRECTED)
plot_edges <- final_dataset_lite %>%
  distinct(unique_pop_id, Range_Position) %>%
  ggplot(aes(x = Range_Position, fill = Range_Position)) +
  # ADDED show.legend = FALSE here
  geom_bar(color = "black", alpha = 0.8, show.legend = FALSE) + 
  scale_fill_manual(values = c("Leading_Edge" = "#4d4d4d", 
                               "Trailing_Edge" = "#999999", 
                               "Centroid" = "#cccccc", 
                               "Optimum" = "#666666")) +
  theme_minimal() +
  labs(title = "C) Populations by range position",
       x = "Measured edge", y = "Number of populations")
  # Removed theme(legend.position = "none") as it is no longer needed
```

Bind
```{r}
attribute_panel <- (plot_eco_type | plot_taxa) / (plot_edges | plot_eco_species) +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")

attribute_panel

ggsave("./stage_4_bind_biotime_bioshifts/Figure_Attribute_Panel.png", 
       plot = attribute_panel, 
       width = 12, height = 10, dpi = 300, bg = "white")
```


##Maps
```{r}
# 1. Convert your unique H3 IDs into exact hexagonal spatial polygons
h3_hexagons <- final_dataset_lite %>%
  distinct(h3_id, Eco)

# Extract raw polygons and convert to sf
h3_hexagons$geometry <- cell_to_polygon(h3_hexagons$h3_id, simple = TRUE)
h3_hexagons <- st_as_sf(h3_hexagons, crs = 4326)

# THE FIX: Split polygons that cross the International Date Line
h3_hexagons <- st_wrap_dateline(h3_hexagons)

world_sf <- ne_countries(scale = "large", type = "map_units", returnclass = "sf")# 1. Americas & Pacific 

```


```{r}
# 1. Americas Map 
plot_map_americas <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), color = "black", 
          alpha = 0.8, linewidth = 0.1) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-14500000, -4500000), ylim = c(1000000, 7000000)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "A) Americas & Pacific", x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    # THE FIX: Negative bottom margin (-20) pulls the title down to the map border
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(t = 10, b = -20)),
    axis.text = element_text(size = 8, color = "grey30"),
    legend.position = c(0.15, 0.2), 
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.title = element_blank(),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

# 2. Europe Focus 
plot_map_europe <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), color = "black", 
          alpha = 0.8, linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-1000000, 2000000), ylim = c(3500000, 7500000)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "B) Europe", x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    # THE FIX: Negative bottom margin
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(t = 10, b = -20)),
    axis.text = element_text(size = 8, color = "grey30"),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

# 3. South Africa Inset 
plot_map_sa <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "grey65", linewidth = 0.3) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), 
          color="black", alpha = 0.8, linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = 4326, xlim = c(15, 35), ylim = c(-40, -22)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "C) South Africa", x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
    panel.grid.major = element_line(color = "grey80", linetype = "dashed", linewidth = 0.3),
    # THE FIX: Negative bottom margin
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(t = 10, b = -20)),
    axis.text = element_text(size = 8, color = "grey30"),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

# Assemble and Save
library(patchwork)
custom_layout <- "
  AAABB
  AAABB
  AAACC
"

final_maps_only <- plot_map_americas + plot_map_europe + plot_map_sa + 
  plot_layout(design = custom_layout)

final_maps_only

ggsave("./stage_4_bind_biotime_bioshifts/maps.png", 
       plot = final_maps_only, 
       width = 14, height = 10, dpi = 300, bg = "white")
```
























```{r}

# 1. Americas Map (Legend inside)
plot_map_americas <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "white", linewidth = 0.01) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), color = "black", 
          alpha = 0.8, linewidth = 0.1) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-14500000, -4500000),
           ylim = c(1000000, 7000000)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "A) Americas & Pacific") +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_blank(), 
    axis.text = element_blank(),
    # DOCK LEGEND INSIDE MAP (adjust x and y from 0 to 1)
    legend.position = c(0.2, 0.2), 
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.title = element_blank()
  )
plot_map_americas

# 2. Europe Focus (No legend)
plot_map_europe <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "white", linewidth = 0.01) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), color = "black", 
          alpha = 0.8, linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = "+proj=robin", 
           xlim = c(-1000000, 2000000),
           ylim = c(3500000, 7500000)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "B) Europe") +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_blank(), 
    axis.text = element_blank()
  )
plot_map_europe
# 3. South Africa Inset (No legend)
plot_map_sa <- ggplot() +
  geom_sf(data = world_sf, fill = "grey90", color = "white", linewidth = 0.01) +
  geom_sf(data = h3_hexagons, aes(fill = Eco), 
          color="black",alpha = 0.8, linewidth = 0.1, show.legend = FALSE) +
  coord_sf(crs = 4326, xlim = c(15, 35), ylim = c(-40, -22)) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f", "Fre" = "#e9c46a")) +
  theme_minimal() +
  labs(title = "C) South Africa") +
  theme(
    panel.background = element_rect(fill = "#e4f1f5", color = NA),
    panel.border = element_blank(), 
    axis.text = element_blank()
  )
plot_map_sa


# THE FIX: Custom Text Design Layout
# 'A' gets 3 columns of width, 'B' and 'C' get 2.
# 'A' and 'B' span 2 rows of height, 'C' gets 1.
custom_layout <- "
  AAABB
  AAABB
  AAACC
"

final_maps_only <- plot_map_americas + plot_map_europe + plot_map_sa + 
  plot_layout(design = custom_layout)

final_maps_only
ggsave("./stage_4_bind_biotime_bioshifts/maps.png", 
       plot = final_maps_only, 
       width = 12, height = 10, dpi = 300, bg = "white")
```
