---
title: "stage_3_stand_biotime"
author: "Shahar Chaikin"
date: "2026-05-19"
output: html_document
---

This script outlines the spatial and temporal standardization pipeline for the BioTIME dataset, preparing raw abundance records for time-series modeling. It defines distinct ecological populations by assigning sampling coordinates to spatial H3 hexagonal grid cells and standardizes intra-annual sampling effort by calculating mean Catch Per Unit Effort (CPUE). Crucially, the pipeline explicitly accounts for implicit non-detections by padding the dataset with "true zeros" derived from the local species pool, while intelligently truncating prolonged trailing zeros to reflect potential local extinctions. Finally, it filters the data to ensure sufficient temporal depth (at least three sampled years) and adequate sampling effort (minimum of four samples per year), yielding a clean, integer-based dataset optimized for count-based mixed-effects modeling.

Libraries
```{r}
library(tidyverse)
library(sf)
library(h3jsr)
```

#1) Load intersect biotime data
```{r}
biotime_intersect=read_rds("./stage_2_intersect_data/biotime_intersect.rds")
```

# 2 Clean NAs and Optimize Spatial H3 Assignment
```{r}
# Extracting unique coordinates is infinitely faster than converting 5.4M rows to sf
unique_coords <- biotime_intersect %>%
  filter(!is.na(ABUNDANCE)) %>%
  select(LONGITUDE, LATITUDE) %>%
  distinct() %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, remove = FALSE)

# Assign H3 cells (Resolution 3) to the unique coordinates
unique_coords$h3_id <- point_to_cell(unique_coords, res = 3)

# Join the H3 IDs back to the main dataset, simultaneously dropping NAs
biotime_spatial <- biotime_intersect %>%
  filter(!is.na(ABUNDANCE)) %>%
  left_join(
    unique_coords %>% st_drop_geometry(), 
    by = c("LONGITUDE", "LATITUDE")
  )

#free RAM
rm(biotime_intersect)
gc()
```

# 3) Define True Zeros (Implicit Non-Detections)
Here we create a grid of all actual sampling events and cross-reference them with the known species pool for each location to expose true zeros.

```{r}
# Get all unique samples taken per study, location, and year
all_samples <- biotime_spatial %>%
  select(STUDY_ID, h3_id, YEAR, SAMPLE_DESC) %>%
  distinct()

# Get the unique "species pool" (and taxonomy) for each location
species_pool <- biotime_spatial %>%
  select(
    STUDY_ID, h3_id, species_harmonized, taxonID, taxon, kingdom, phylum, class, order, family
  ) %>%
  distinct()

# Cross-reference samples with the species pool to create the expected observation grid
full_grid <- all_samples %>%
  left_join(species_pool, by = c("STUDY_ID", "h3_id"), relationship = "many-to-many")

# CRITICAL FIX: Pre-aggregate the raw abundances to sum up any split records in the same sample
actual_abundances <- biotime_spatial %>%
  group_by(SAMPLE_DESC, species_harmonized) %>%
  summarise(ABUNDANCE = sum(ABUNDANCE, na.rm = TRUE), .groups = "drop")

# Join the aggregated abundances into the grid, replacing NAs with 0
biotime_zeros <- full_grid %>%
  left_join(
    actual_abundances,
    by = c("SAMPLE_DESC", "species_harmonized")
  ) %>%
  mutate(ABUNDANCE = replace_na(ABUNDANCE, 0))

# Free up RAM
rm(all_samples, species_pool, full_grid, actual_abundances, biotime_spatial)
gc()
```

#4) Standardize Effort & Calculate Mean Abundance
Filter out years with insufficient sampling effort (minimum 4 samples) and calculate the standardized Catch Per Unit Effort (CPUE).

```{r}
biotime_ts <- biotime_zeros %>%
  group_by(
    STUDY_ID, h3_id, YEAR, species_harmonized, taxonID, taxon, kingdom, phylum, class, order, family
  ) %>%
  summarise(
    mean_annual_abundance = mean(ABUNDANCE),
    n_samples_available = n(),
    .groups = "drop"
  ) %>%
  # Apply the minimum effort control
  filter(n_samples_available >= 4) %>%
  # Round to integer for downstream count-based mixed models (Poisson/NB)
  mutate(mean_annual_abundance = round(mean_annual_abundance)) %>%
  # Create the unique population identifier
  mutate(
    unique_pop_id = paste(STUDY_ID, h3_id, species_harmonized, sep = "_")
  ) %>%
  relocate(unique_pop_id)

# Free up RAM
rm(biotime_zeros)
gc()
```

#5) Truncate Trailing Zeros
If a species disappears from a location and yields more than 4 consecutive years of zeros at the end of its time series, we assume true absence/local extinction and truncate those trailing zeros.
```{r}
biotime_ts_final <- biotime_ts %>%
  group_by(unique_pop_id) %>%
  arrange(YEAR) %>%
  mutate(
    # Check if the species was ever present (safety check)
    ever_present = any(mean_annual_abundance > 0),
    
    # Find the row index of the very last year it was detected
    last_presence_idx = max(which(mean_annual_abundance > 0)),
    
    # Identify which rows are trailing zeros (occur after the last presence)
    is_trailing_zero = row_number() > last_presence_idx,
    
    # Count the total number of trailing zeros at the end of the time series
    total_trailing_zeros = sum(is_trailing_zero)
  ) %>%
  # Keep the row IF it is NOT a trailing zero, OR if the total trailing run is <= 4
  filter(ever_present & (!is_trailing_zero | total_trailing_zeros <= 4)) %>%
  # Clean up temporary calculation columns
  select(-ever_present, -last_presence_idx, -is_trailing_zero, -total_trailing_zeros) %>%
  ungroup()

# Free up RAM
rm(biotime_ts)
gc()

# Final structure check
head(biotime_ts_final)
```

# 6) Minimum Time-Series Length Filter
Keep only populations that have been sampled in at least 3 distinct years.

```{r}
biotime_ts_final <- biotime_ts_final %>%
  group_by(unique_pop_id) %>%
  filter(n() >= 3) %>%
  ungroup()

# Quick summary of the final dataset size
cat("Final number of unique populations:", n_distinct(biotime_ts_final$unique_pop_id), "\n")
cat("Total rows ready for modeling:", nrow(biotime_ts_final), "\n")
```

#7) Save Standardized Time Series
```{r}
# Save the final dataset
saveRDS(biotime_ts_final, "./stage_3_std_biotime_intersect/biotime_ts_final.rds")
```