---
title: "stage_5_models"
author: "Shahar Chaikin"
date: "2026-05-19"
output: html_document
---
```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
```

## 1. Environment Setup

Load the necessary statistical and data manipulation packages.

```{r libraries}
# Data manipulation and visualization
library(tidyverse)

# Advanced mixed modeling
library(glmmTMB)

# Diagnostics and performance
library(DHARMa)       # Gold standard for mixed model residuals
library(performance)  # Comprehensive model checks

# Tidy coefficient extraction and R-squared
library(broom.mixed)  # Tidyverse support for glmmTMB
library(MuMIn)        # Pseudo R-squared for mixed models
library(ggeffects)
```

## 2. Data Loading and Preparation

Here, we load the harmonized dataset and isolate the latitudinal shifts. We also scale our continuous predictors (`YEAR` and `Rate`) to help the model converge faster and make the interaction coefficients directly comparable.

```{r data_prep}
# 1. Load the Thermal Position Index
thermal_data <- read_rds("./stage_6_thermal_position/thermal_pos_all_ecos.rds") %>%
  dplyr::select(unique_pop_id, thermal_pos_index)

# 2. Prepare and properly format the dataset for the model
dataset_for_model <- read_rds("./stage_4_bind_biotime_bioshifts/master_population_slopes.rds") %>%
  # STRICT FILTER: Ensure we are only comparing Latitudinal shifts
  filter(toupper(Type) == "LAT") %>%
  
  # THE FIX: Join the thermal position index
  left_join(thermal_data, by = "unique_pop_id") %>%
  drop_na(thermal_pos_index) %>% # Essential: Drop NAs before feeding to the model
  
  # Format factors for clean modeling and plotting
  mutate(
    # Set Terrestrial as the baseline for the interaction contrast
    Eco = factor(Eco, levels = c("Ter", "Mar")),
    
    # Cast grouping variables as factors so they behave correctly as random intercepts
    species_harmonized = as.factor(species_harmonized),
    h3_id = as.factor(h3_id),
    model_type = as.factor(model_type),
    class = as.factor(class),
    ID_BioShifts = as.factor(ID_BioShifts),
   
    # Find the 99th percentile of the weights
    weight_99th = quantile(inv_var_weight, 0.99, na.rm = TRUE),
    
    # Cap any weight above that threshold strictly at the 99th percentile
    weight_capped = pmin(inv_var_weight, weight_99th),
    weight_years = log(total_sampled_years)
  ) %>% 
  ungroup()


```

## 3.Modeling

###3.1 run glmm
The effect of range shift velocities over time and realms on abundance

```{r}
final_eco_model <- glmmTMB(
  
  slope ~ scale(Rate) * Eco * scale(thermal_pos_index) + 
    
    (1 | species_harmonized) + 
    (1 | h3_id) + 
    (1 | ID_BioShifts),
  
  data = dataset_for_model,
  weights = weight_years, 
  family = t_family(),
  
  # THE FIX: Properly formatted control arguments to give the optimizer 10x more steps
  control = glmmTMBControl(
    optCtrl = list(iter.max = 10000, eval.max = 10000)
  )
)
#diagnostics
check_convergence(final_eco_model)
check_collinearity(final_eco_model)
check_singularity(final_eco_model)
sim_res <- simulateResiduals(fittedModel = final_eco_model, plot = TRUE,n = 70)


saveRDS(final_eco_model,"./stage_7_models/final_eco_model_t_pos.rds")
summary(final_eco_model)
final_eco_model_rsq = r.squaredGLMM(final_eco_model)
final_eco_model_rsq


# 1. Extract the Pearson residuals
diag_data <- data.frame(
  resid = residuals(final_eco_model, type = "pearson")
) %>% na.omit()

# 2. Open the PNG pipeline
png("./stage_7_models/QQ_Plot_Student_t_Zoomed.png", width = 6, height = 6, units = "in", res = 300)

# 3. Generate theoretical quantiles 
theoretical_t <- qt(ppoints(length(diag_data$resid)), df = 1.7)

# 4. Draw the zoomed Q-Q plot
qqplot(theoretical_t, diag_data$resid, 
       pch = 16, 
       col = rgb(0, 0, 0, alpha = 0.1), 
       main = "Zoomed Student-t Q-Q Plot",
       xlab = "Theoretical Quantiles (Student-t, df = 1.7)",
       ylab = "Sample Quantiles (Weighted Pearson Residuals)",
       
       # THE FIX: Zoom the camera in to the central data cloud!
       xlim = c(-15, 15), 
       ylim = c(-350, 350)) 

# 5. Add the reference line
probs <- c(0.25, 0.75)
y_coords <- quantile(diag_data$resid, probs, names = FALSE, na.rm = TRUE)
x_coords <- qt(probs, df = 1.7)
slope <- diff(y_coords) / diff(x_coords)
int <- y_coords[1] - slope * x_coords[1]
abline(int, slope, col = "red", lwd = 2, lty = 2)

# 6. Close the pipeline
dev.off()

cat("Zoomed QQ Plot saved!\n")

sjPlot::plot_model(final_eco_model,type="pred",terms=c("Rate","thermal_pos_index[-1,1]","Eco"))
```

###3.2 summary table
```{r}
# Extract the fixed effects cleanly using broom.mixed
model_coefficients <- tidy(final_eco_model, effects = "fixed", conf.int = TRUE) %>%
  # Round numeric columns for clean reading
  mutate(across(where(is.numeric), ~ round(., 4)))

# Export the coefficient table
write_csv(model_coefficients, "./stage_7_models/Table_1_Fixed_Effects_t_pos.csv")
cat("Exported Fixed Effects table to CSV.\n")

# Format the R-squared values into a clean data frame
r2_table <- data.frame(
  Variance_Explained = c("Marginal (Fixed Effects)", "Conditional (Fixed + Random)"),
  R_Squared = c(round(final_eco_model_rsq[1, "R2m"], 4), 
                round(final_eco_model_rsq[1, "R2c"], 4))
)

# Export the R-squared table
write_csv(r2_table, "./stage_7_models/Table_2_R_Squared_t_pos.csv")
cat("Exported R-Squared table to CSV.\n")
```

###3.3 Plot
```{r}
# 1. Extract marginal effects for the interaction
# ggeffects automatically handles the scale() back-transformation
predicted_trends <- ggpredict(final_eco_model, terms = c("Rate [all]", "thermal_pos_index[-1,1]","Eco"))

# 2. Build the plot
plot_final_interaction <- ggplot(predicted_trends, aes(x = x, y = predicted, color = group, fill = group)) +
  
  # Add the confidence ribbons (no borders for a cleaner look)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  
  # Add the predicted trend lines
  geom_line(linewidth = 1.2) +
  
  # Reference lines for Zero Velocity (x=0) and Stable Population (y=0)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # THE FIX: Added y = slope and sides = "bl" to show density on both axes
  geom_rug(data = dataset_for_model, 
           aes(x = Rate, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
  
  # Match the aesthetic styling perfectly
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  # TRUNCATE THE AXES to show the biological reality, preventing the CIs 
  # from flaring out endlessly at the extreme, data-poor boundaries.
  # Note: Adjust the xlim and ylim if your data density sits slightly wider!
  coord_cartesian(xlim = c(-29, 21), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Predicted temporal trend (annual slope)",
    color = "Realm",
    fill = "Realm"
  ) +
  theme(
    legend.position = c(0.85, 0.85),
    legend.background = element_rect(fill = "white", color = "grey60", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11)
  )

# Print the plot in the document
print(plot_final_interaction)

# Save for the manuscript
ggsave("./stage_7_models/Final_Interaction_Plot.png", 
       plot = plot_final_interaction, width = 8, height = 6, dpi = 300)
```


Plot on original scale
```{r}
# 1. Calculate exact sample sizes for the caption
n_mar <- sum(dataset_for_model$Eco == "Mar", na.rm = TRUE)
n_ter <- sum(dataset_for_model$Eco == "Ter", na.rm = TRUE)

# 2. Create a dynamic, formatted caption string with a line break (\n)
r2_caption <- sprintf(
  "Marginal R² = %.3f | Conditional R² = %.3f\nPopulations modeled (n): %d Marine | %d Terrestrial", 
  final_eco_model_rsq[1, "R2m"], 
  final_eco_model_rsq[1, "R2c"],
  n_mar, 
  n_ter
)

# 3. Extract marginal effects for the 3-way interaction
predicted_trends <- ggpredict(final_eco_model, terms = c("Rate [all]", "Eco", "thermal_pos_index [-1, 0, 1]")) %>%
  as.data.frame() %>%
  mutate(facet = factor(facet, 
                        levels = c("-1", "0", "1"), 
                        labels = c("Cold edge (-1)", "Core range (0)", "Warm edge (+1)")))

# 4. Build the Raw Scale Plot
plot_final_interaction <- ggplot(predicted_trends, aes(x = x, y = predicted, color = group, fill = group, group = group)) +
  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  geom_rug(data = dataset_for_model, 
           aes(x = Rate, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
           
  facet_wrap(~facet) +
  
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  coord_cartesian(xlim = c(-29, 21), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Predicted abundance trend (slope)",
    color = "Realm",
    fill = "Realm",
    caption = r2_caption # Dynamically populated with R² and sample sizes
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90", color = "grey20"),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 10, face = "italic", color = "grey40", hjust = 1) 
  )

# Print and Save
ggsave("./stage_7_models/Final_Interaction_Plot_Faceted.png", 
       plot = plot_final_interaction, width = 10, height = 5, dpi = 300)
```

# 2. Build the plot
# THE FIX: Added `group = group` explicitly to force the lines to draw
plot_final_interaction <- ggplot(predicted_trends, aes(x = x, y = predicted, color = group, fill = group, group = group)) +
  
  # Add the confidence ribbons (no borders for a cleaner look)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  
  # Add the predicted trend lines
  geom_line(linewidth = 1.2) +
  
  # Reference lines for Zero Velocity (x=0) and Stable Population (y=0)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # Rug marks to show density
  geom_rug(data = dataset_for_model, 
           aes(x = Rate, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
           
  # Create side-by-side panels for the thermal positions
  facet_wrap(~facet) +
  
  # Match the aesthetic styling perfectly
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  # TRUNCATE THE AXES to show the biological reality
  coord_cartesian(xlim = c(-29, 21), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Predicted temporal trend (annual slope)",
    color = "Realm",
    fill = "Realm"
  ) +
  theme(
    legend.position = "bottom", # Moved legend to bottom to accommodate the 3 panels
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90", color = "grey20"),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11)
  )

# Print the plot in the document
print(plot_final_interaction)

# Save for the manuscript
ggsave("./stage_7_models/Final_Interaction_Plot_Faceted.png", 
       plot = plot_final_interaction, width = 10, height = 5, dpi = 300)
```
