---
title: "stage_7_models"
author: "Shahar Chaikin"
date: "2026-05-19"
output: html_document
---
```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
```

## 1. Environment Setup

Load the necessary statistical and data manipulation packages.

```{r libraries}
# Data manipulation and visualization
library(tidyverse)

# Advanced mixed modeling
library(glmmTMB)

# Diagnostics and performance
library(DHARMa)       # Gold standard for mixed model residuals
library(performance)  # Comprehensive model checks

# Tidy coefficient extraction and R-squared
library(broom.mixed)  # Tidyverse support for glmmTMB
library(MuMIn)        # Pseudo R-squared for mixed models
library(ggeffects)
```

## 2. Data Loading and Preparation

Here, we load the harmonized dataset and isolate the latitudinal shifts. We also scale our continuous predictors (`YEAR` and `Rate`) to help the model converge faster and make the interaction coefficients directly comparable.

```{r data_prep}
# 1. Load the Thermal Position Index
thermal_data <- read_rds("./stage_6_thermal_position/thermal_pos_all_ecos.rds") %>%
  dplyr::select(unique_pop_id, thermal_pos_index)

# 2. Prepare and properly format the dataset for the model
dataset_for_model <- read_rds("./stage_4_bind_biotime_bioshifts/master_population_slopes.rds") %>%
  # STRICT FILTER: Ensure we are only comparing Latitudinal shifts
  filter(toupper(Type) == "LAT") %>%
  
  # THE FIX: Join the thermal position index
  left_join(thermal_data, by = "unique_pop_id") %>%
  drop_na(thermal_pos_index) %>% # Essential: Drop NAs before feeding to the model
  
  # Format factors for clean modeling and plotting
  mutate(
    # Set Terrestrial as the baseline for the interaction contrast
    Eco = factor(Eco, levels = c("Ter", "Mar")),
    
    # Cast grouping variables as factors so they behave correctly as random intercepts
    species_harmonized = as.factor(species_harmonized),
    h3_id = as.factor(h3_id),
    model_type = as.factor(model_type),
    class = as.factor(class),
    ID_BioShifts = as.factor(ID_BioShifts),
   
    # Find the 99th percentile of the weights
    weight_99th = quantile(inv_var_weight, 0.99, na.rm = TRUE),
    
    # Cap any weight above that threshold strictly at the 99th percentile
    weight_capped = pmin(inv_var_weight, weight_99th),
    weight_years = log(total_sampled_years)
  ) %>% 
  ungroup()

#write_rds(dataset_for_model,"./stage_7_models/dataset_for_model.rds")
```

## 3.Modeling

###3.1 run glmm
The effect of range shift velocities over time and realms on abundance

```{r}
final_eco_model <- glmmTMB(
  
  slope ~ scale(Rate) * Eco * scale(thermal_pos_index) + 
    
    (1 | species_harmonized) + 
    (1 | h3_id) + 
    (1 | ID_BioShifts),
  
  data = dataset_for_model,
  weights = weight_years, 
  family = t_family(),
  
  # THE FIX: Properly formatted control arguments to give the optimizer 10x more steps
  control = glmmTMBControl(
    optCtrl = list(iter.max = 10000, eval.max = 10000)
  )
)
#diagnostics
check_convergence(final_eco_model)
check_collinearity(final_eco_model)
check_singularity(final_eco_model)
sim_res <- simulateResiduals(fittedModel = final_eco_model, plot = TRUE,n = 70)


#saveRDS(final_eco_model,"./stage_7_models/final_eco_model_t_pos.rds")
summary(final_eco_model)
final_eco_model_rsq = r.squaredGLMM(final_eco_model)
final_eco_model_rsq


# 1. Extract the Pearson residuals
diag_data <- data.frame(
  resid = residuals(final_eco_model, type = "pearson")
) %>% na.omit()

# 2. Open the PNG pipeline
png("./stage_7_models/QQ_Plot_Student_t_Zoomed.png", width = 6, height = 6, units = "in", res = 300)

# 3. Generate theoretical quantiles 
theoretical_t <- qt(ppoints(length(diag_data$resid)), df = 1.7)

# 4. Draw the zoomed Q-Q plot
qqplot(theoretical_t, diag_data$resid, 
       pch = 16, 
       col = rgb(0, 0, 0, alpha = 0.1), 
       main = "Zoomed Student-t Q-Q Plot",
       xlab = "Theoretical Quantiles (Student-t, df = 1.7)",
       ylab = "Sample Quantiles (Weighted Pearson Residuals)",
       
       # THE FIX: Zoom the camera in to the central data cloud!
       xlim = c(-15, 15), 
       ylim = c(-350, 350)) 

# 5. Add the reference line
probs <- c(0.25, 0.75)
y_coords <- quantile(diag_data$resid, probs, names = FALSE, na.rm = TRUE)
x_coords <- qt(probs, df = 1.7)
slope <- diff(y_coords) / diff(x_coords)
int <- y_coords[1] - slope * x_coords[1]
abline(int, slope, col = "red", lwd = 2, lty = 2)

# 6. Close the pipeline
dev.off()

cat("Zoomed QQ Plot saved!\n")

sjPlot::plot_model(final_eco_model,type="pred",terms=c("Rate","thermal_pos_index[-1,1]","Eco"))
```

###3.2 summary table
```{r}
# Extract the fixed effects cleanly using broom.mixed
model_coefficients <- tidy(final_eco_model, effects = "fixed", conf.int = TRUE) %>%
  # Round numeric columns for clean reading
  mutate(across(where(is.numeric), ~ round(., 4)))

# Export the coefficient table
write_csv(model_coefficients, "./stage_7_models/Table_1_Fixed_Effects_t_pos.csv")
cat("Exported Fixed Effects table to CSV.\n")

# Format the R-squared values into a clean data frame
r2_table <- data.frame(
  Variance_Explained = c("Marginal (Fixed Effects)", "Conditional (Fixed + Random)"),
  R_Squared = c(round(final_eco_model_rsq[1, "R2m"], 4), 
                round(final_eco_model_rsq[1, "R2c"], 4))
)

# Export the R-squared table
write_csv(r2_table, "./stage_7_models/Table_2_R_Squared_t_pos.csv")
cat("Exported R-Squared table to CSV.\n")
```

###3.3 Plot
```{r}
# 1. Extract marginal effects for the interaction
# ggeffects automatically handles the scale() back-transformation
predicted_trends <- ggpredict(final_eco_model, terms = c("Rate [all]", "thermal_pos_index[-1,1]","Eco"))

# 2. Build the plot
plot_final_interaction <- ggplot(predicted_trends, aes(x = x, y = predicted, color = group, fill = group)) +
  
  # Add the confidence ribbons (no borders for a cleaner look)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  
  # Add the predicted trend lines
  geom_line(linewidth = 1.2) +
  
  # Reference lines for Zero Velocity (x=0) and Stable Population (y=0)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # THE FIX: Added y = slope and sides = "bl" to show density on both axes
  geom_rug(data = dataset_for_model, 
           aes(x = Rate, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
  
  # Match the aesthetic styling perfectly
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  # TRUNCATE THE AXES to show the biological reality, preventing the CIs 
  # from flaring out endlessly at the extreme, data-poor boundaries.
  # Note: Adjust the xlim and ylim if your data density sits slightly wider!
  coord_cartesian(xlim = c(-29, 21), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Predicted temporal trend (annual slope)",
    color = "Realm",
    fill = "Realm"
  ) +
  theme(
    legend.position = c(0.85, 0.85),
    legend.background = element_rect(fill = "white", color = "grey60", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11)
  )

# Print the plot in the document
print(plot_final_interaction)

# Save for the manuscript
ggsave("./stage_7_models/Final_Interaction_Plot.png", 
       plot = plot_final_interaction, width = 8, height = 6, dpi = 300)
```

Plot on original scale
```{r}
# 1. Extract marginal effects
predicted_trends_orig <- ggpredict(final_eco_model, terms = c("Rate [all]", "Eco", "thermal_pos_index [-1, 0, 1]")) %>%
  as.data.frame() %>%
  mutate(facet = factor(facet, 
                        levels = c("-1", "0", "1"), 
                        labels = c("Cold Edge (-1)", "Core Range (0)", "Warm Edge (+1)")))

# Generate the exact mathematical positions for 20% jumps
target_breaks <- log(1 + seq(-0.8, 1.8, by = 0.2))

# 2. Build the Percentage Scale Plot
plot_final_interaction_orig <- ggplot(predicted_trends_orig, aes(x = x, y = predicted, color = group, fill = group, group = group)) +
  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  geom_rug(data = dataset_for_model, 
           aes(x = Rate, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
           
  facet_wrap(~facet) +
  
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  scale_y_continuous(
    breaks = target_breaks,
    labels = function(x) paste0(round((exp(x) - 1) * 100, 1), "%")
  ) +
  
  coord_cartesian(xlim = c(-15, 15), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Annual population change (%)", 
    color = "Realm",
    fill = "Realm",
    caption = r2_caption # Added the exact same dynamic caption here
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90", color = "grey20"),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 10, face = "italic", color = "grey40", hjust = 1) 
  )

# Print and Save
ggsave("./stage_7_models/Final_Interaction_Plot2_Faceted.png", 
       plot = plot_final_interaction_orig, width = 10, height = 5, dpi = 300)
```

slopes~thermal position
```{r}
# 1. Extract marginal effects (X = Thermal Position, Color = Eco)
# By omitting 'Rate' from terms, it is automatically held at its mean (0 km/yr)
predicted_pure_thermal <- ggpredict(final_eco_model, terms = c("thermal_pos_index [all]", "Eco")) %>%
  as.data.frame()

# 2. Re-calculate caption 
n_mar <- sum(dataset_for_model$Eco == "Mar", na.rm = TRUE)
n_ter <- sum(dataset_for_model$Eco == "Ter", na.rm = TRUE)
r2_caption <- sprintf(
  "Marginal R² = %.3f | Conditional R² = %.3f\nPopulations modeled (n): %d Marine | %d Terrestrial", 
  final_eco_model_rsq[1, "R2m"], final_eco_model_rsq[1, "R2c"], n_mar, n_ter
)

# 3. Build the Raw Scale Plot (No Facets)
plot_pure_thermal <- ggplot(predicted_pure_thermal, aes(x = x, y = predicted, color = group, fill = group, group = group)) +
  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  
  # Y=0 is a stable population. X=0 is the thermal core of the species' range.
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # Rug marks mapping the density of the thermal positions
  geom_rug(data = dataset_for_model, 
           aes(x = thermal_pos_index, y = slope, color = Eco), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
  
  scale_color_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                     labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  scale_fill_manual(values = c("Mar" = "#0077b6", "Ter" = "#2a9d8f"),
                    labels = c("Mar" = "Marine", "Ter" = "Terrestrial")) +
  
  # Truncating the X-axis to the biological bulk of the thermal index 
  coord_cartesian(xlim = c(-1.5, 1.5), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    title = "Overall population trends across the thermal niche",
    x = "Thermal Position Index (Cold Edge <--- 0 ---> Warm Edge)",
    y = "Predicted temporal trend (annual slope)",
    color = "Realm",
    fill = "Realm",
    caption = r2_caption
  ) +
  theme(
    legend.position = c(0.85, 0.85), # Moved legend inside the plot area for a single panel
    legend.background = element_rect(fill = "white", color = "grey60", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 10, face = "italic", color = "grey40", hjust = 1) 
  )

ggsave("./stage_7_models/Pure_Thermal_Perspective_Plot.png", 
       plot = plot_pure_thermal, width = 8, height = 6, dpi = 300)
```

#4) class level
```{r}
# 1. Determine which classes have enough statistical power
# Must have >50 populations AND at least 5 distinct species
valid_classes <- dataset_for_model %>%
  group_by(class) %>%
  summarise(n_pops = n(), n_species = n_distinct(species_harmonized)) %>%
  filter(n_pops >= 100, n_species >= 5) %>%
  pull(class)

# 2. Filter, drop unused levels, and PRE-SCALE variables safely
class_dataset <- dataset_for_model %>%
  filter(class %in% valid_classes) %>%
  mutate(
    class = droplevels(class),
    # Pre-scaling prevents Hessian calculation errors in glmmTMB
    Rate_scaled = as.numeric(scale(Rate)),
    thermal_pos_scaled = as.numeric(scale(thermal_pos_index))
  )

cat("Modeling across", n_distinct(class_dataset$class), "robust classes:\n")
print(table(class_dataset$class))
```

##4.1) Fit the Class-Level Model (Realm removed)
```{r class_model}
class_only_model <- glmmTMB(
  
  # A pure 3-way interaction to test taxonomic differences
 slope ~ Rate_scaled * class * thermal_pos_scaled + 
    
    (1 | species_harmonized) + 
    (1 | h3_id), 
    # (1 | ID_BioShifts) is dropped to prevent confounding
  
  data = class_dataset,
  
  # Scaled weights to prevent variance collapse
  weights = weight_years, 
  
  family = gaussian(),
  
  # THE FIX 2: Correct syntax for passing arguments to the BFGS optimizer
  control = glmmTMBControl(
    optimizer = optim,
    optArgs = list(method = "BFGS"),    # The engine
    optCtrl = list(maxit = 10000)       # The max iterations
  )
)

#diagnostics
check_convergence(class_only_model)
check_collinearity(class_only_model)
check_singularity(class_only_model)
sim_res <- simulateResiduals(fittedModel = class_only_model, plot = TRUE,n = 70)


saveRDS(class_only_model, "./stage_7_models/class_only_model_t_pos.rds")

summary(class_only_model)
class_model_rsq <- r.squaredGLMM(class_only_model)
print(class_model_rsq)
```

```{r class_plot, fig.width=12, fig.height=6}
## 4. Plotting Class-Level Predictions (Gaussian Model)
library(ggeffects)

# 1. Extract marginal effects using the PRE-SCALED variable names
predicted_class_trends <- ggpredict(class_only_model, 
                                    terms = c("Rate_scaled [all]", "class", "thermal_pos_scaled [-1, 0, 1]")) %>%
  as.data.frame() %>%
  mutate(facet = factor(facet, 
                        levels = c("-1", "0", "1"), 
                        labels = c("Cold Edge (-1)", "Core Range (0)", "Warm Edge (+1)")))

# 2. Setup dynamic caption
class_model_rsq <- MuMIn::r.squaredGLMM(class_only_model)
r2_caption_class <- sprintf(
  "Marginal R² = %.3f | Conditional R² = %.3f\nPopulations modeled (n): %d across %d classes", 
  class_model_rsq[1, "R2m"], class_model_rsq[1, "R2c"], 
  nrow(class_dataset), n_distinct(class_dataset$class)
)

# 3. Define breaks for percentage scale (assuming slope is log-ratio)
target_breaks <- log(1 + seq(-0.8, 1.8, by = 0.2))

# 4. Build the Plot
plot_class_interaction <- ggplot(predicted_class_trends, aes(x = x, y = predicted, color = group, fill = group, group = group)) +
  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.2) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # Rug marks matched to class (using scaled variables to match the x-axis)
  geom_rug(data = class_dataset, 
           aes(x = Rate_scaled, y = slope, color = class), 
           sides = "bl", alpha = 0.3, inherit.aes = FALSE) +
           
  facet_wrap(~facet) +
  
  scale_color_viridis_d(option = "turbo", end = 0.9) +
  scale_fill_viridis_d(option = "turbo", end = 0.9) +
  
  scale_y_continuous(
    breaks = target_breaks,
    labels = function(x) paste0(round((exp(x) - 1) * 100, 1), "%")
  ) +
  
  # Note: Because the X-axis is now scaled, we adjust the xlim constraints
  coord_cartesian(xlim = c(-3, 3), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    x = "Range shift velocity (Scaled)",
    y = "Annual population change (%)", 
    color = "Taxonomic Class",
    fill = "Taxonomic Class",
    caption = r2_caption_class
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90", color = "grey20"),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 10, face = "italic", color = "grey40", hjust = 1) 
  )

print(plot_class_interaction)
ggsave("./stage_7_models/Class_Interaction_Plot_Faceted.png", 
       plot = plot_class_interaction, width = 12, height = 5, dpi = 300)
```

#5) Chaikin et et al., 2024 populations

Extract hexagons
```{r}
library(sf)
library(h3jsr)

chaikin_2020_pop=read.csv("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\25006787\\data.csv")

# 1. Map the 2024 paper populations to H3 Resolution 3
chaikin_2024_mapped <- chaikin_2020_pop %>%
  # Filter out any populations missing coordinates just to be safe
  filter(!is.na(longitude_mean) & !is.na(latitude_mean)) %>%
  
  # Convert the dataframe to an sf spatial object
  # remove = FALSE ensures your original lat/lon columns stay in the dataframe
  st_as_sf(coords = c("longitude_mean", "latitude_mean"), crs = 4326, remove = FALSE) %>%
  
  # Extract the H3 resolution 3 hexagon ID based on the point geometry
  mutate(h3_id = point_to_cell(geometry, res = 3)) %>%
  
  # Drop the sf geometry column to return it to a clean, standard dataframe
  st_drop_geometry()

# Quick check to verify the new column
cat("Mapped", nrow(chaikin_2024_mapped), "populations to", n_distinct(chaikin_2024_mapped$h3_id), "unique H3 hexagons.\n")

head(chaikin_2024_mapped %>% select(unique_id, species, latitude_mean, longitude_mean, h3_id))
```

Subset the recent data
```{r}
# 1. Subset the new dataset using the spatial and taxonomic footprints of the old dataset
chaikin_2024_subset <- dataset_for_model %>%
  semi_join(chaikin_2024_mapped, 
            by = c("h3_id" = "h3_id", "species_harmonized" = "species")) %>%
  # Ensure the factor levels are clean after subsetting
  mutate(
    species_harmonized = droplevels(species_harmonized),
    h3_id = droplevels(h3_id)
  )

# Let's see how many populations survived the crosswalk
cat("Original 2024 populations:", nrow(chaikin_2024_mapped), "\n")
cat("Matched populations in new synthesis:", nrow(chaikin_2024_subset), "\n")
cat("Proportion retained:", round((nrow(chaikin_2024_subset) / nrow(chaikin_2024_mapped)) * 100, 1), "%\n")
cat("Unique matched species:", n_distinct(chaikin_2024_subset$species_harmonized), "\n")
```

```{r}
# 2. Fit the 2-way sensitivity model
sensitivity_model <- glmmTMB(
  
  # The 2-way interaction: Does velocity impact abundance differently depending on thermal edge?
  slope ~ scale(Rate) * scale(thermal_pos_index) + 
    
    (1 | species_harmonized) + 
    (1 | h3_id), 
    # Note: If this crashes, drop (1 | ID_BioShifts) if your old data only spans 1 or 2 studies
  
  data = chaikin_2024_subset,
  weights = weight_years, 
  family = gaussian(),
  
  control = glmmTMBControl(
    optCtrl = list(iter.max = 10000, eval.max = 10000)
  )
)

# Diagnostics and Summary
check_convergence(sensitivity_model)
summary(sensitivity_model)
```

```{r}
library(ggeffects)

# Extract marginal effects
predicted_sens <- ggpredict(sensitivity_model, 
                            terms = c("Rate [all]", "thermal_pos_index [-1, 0, 1]")) %>%
  as.data.frame() %>%
  mutate(facet = factor(group, 
                        levels = c("-1", "0", "1"), 
                        labels = c("Cold Edge (-1)", "Core Range (0)", "Warm Edge (+1)")))

# Define breaks for percentage scale
target_breaks <- log(1 + seq(-0.8, 1.8, by = 0.2))

# Build the Plot
plot_sensitivity <- ggplot(predicted_sens, aes(x = x, y = predicted)) +
  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "#0077b6") +
  geom_line(color = "#0077b6", linewidth = 1.2) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20", linewidth = 0.8) +
  
  # Add the raw data as background points to show the subset footprint
  geom_point(data = chaikin_2024_subset, 
             aes(x = Rate, y = slope), 
             alpha = 0.1, color = "grey40", size = 1, inherit.aes = FALSE) +
             
  facet_wrap(~facet) +
  
  scale_y_continuous(
    breaks = target_breaks,
    labels = function(x) paste0(round((exp(x) - 1) * 100, 1), "%")
  ) +
  
  coord_cartesian(xlim = c(-15, 15), ylim = c(-1, 1)) + 
  
  theme_bw() +
  labs(
    title = "Sensitivity Analysis: Chaikin et al. (2024) Baseline",
    subtitle = paste("n =", nrow(chaikin_2024_subset), "matched populations"),
    x = expression("Range shift velocity (" * km ~ yr^-1 * ")"),
    y = "Annual population change (%)"
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90", color = "grey20"),
    plot.title = element_text(face = "bold", size = 14)
  )

print(plot_sensitivity)
```

