---
title: "Estimating Ecosystem-Aware Thermal Position"
author: "Shahar Chaikin"
date: "`r Sys.Date()`"
output: html_document
---

Aim:
This script estimates the "thermal position" of populations across Marine, Terrestrial, and Freshwater ecosystems by matching species occurrences and H3 hexagonal cells against the appropriate ecosystem-specific climate raster.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)    
library(raster)       # For legacy raster handling (though terra is recommended for the future)
library(h3jsr)        
library(sf)           
library(DataExplorer)
library(stars)
library(geodata)
library(terra)
```

1) Load spatial thinned occurrence data and Ecosystem Metadata
We must join the Eco metadata so the script knows which raster to use for each species.
```{r}
# 1) Load spatial thinned occurrence data and Ecosystem Metadata

# 1. Load the target species list and FORCE exactly one Ecosystem per species
sp_metadata <- read.csv("./stage_4_bind_biotime_bioshifts/target_species_for_occurrences.csv") %>%
  dplyr::select(species_harmonized, Eco) %>%
  drop_na(Eco) %>% # Drop any rows without an ecosystem first
  distinct(species_harmonized, .keep_all = TRUE) # THE FIX: Strictly 1 row per species

# 2. Load the master cleaned occurrences just gathered
master_occ <- read_rds("./stage_5_coords_ext/master_cleaned_occurrences.rds") %>%
  # Join the ecosystem type so the script knows which raster to use
  left_join(sp_metadata, by = c("species" = "species_harmonized")) %>%
  drop_na(Eco) # Ensure every record has an ecosystem assigned

# Quick sanity check: This should now return 0!
cat("Duplicated species in lookup table:", sum(duplicated(sp_metadata$species_harmonized)), "\n")
```

2) Load Dual Climate Rasters

Marine
```{r}
# ==========================================
# A. MARINE RASTER (GLORYS 1993–2016 SST)
# ==========================================
nc_filepath <- "./stage_6_thermal_position/cmems_mod_glo_phy_my_0.083deg-climatology_P1M-m_1780305040080.nc" 

cat("Loading marine file directly from disk...\n")
clim_data <- read_stars(nc_filepath)

# Collapse the dimensions (averaging across months and depth)
cat("Averaging across months and depth...\n")
annual_mean_stars <- st_apply(clim_data, c("x", "y"), mean, na.rm = TRUE)

# Convert to legacy Raster
mar_raster <- as(annual_mean_stars, "Raster")
raster::crs(mar_raster) <- "EPSG:4326"

# THE FIX: Explicitly name the marine layer so extraction is predictable
names(mar_raster) <- "mar_temp"

cat("SST Raster successfully prepared!\n")
```

Terrestrial
```{r}
# ==========================================
# B. TERRESTRIAL RASTER (WorldClim Annual Mean Temp)
# ==========================================
cat("Fetching WorldClim Terrestrial Raster...\n")

wc_data <- worldclim_global(var = "bio", res = 5, path = "./stage_6_thermal_position")

ter_raster_spat <- wc_data[["wc2.1_5m_bio_1"]]
ter_raster <- raster::raster(ter_raster_spat)

# THE FIX: Explicitly name the terrestrial layer
names(ter_raster) <- "ter_temp"

cat("Terrestrial Raster successfully prepared!\n")
```

View
```{r}
# Define a smooth temperature color palette (cool = blue, warm = red)
temp_pal <- colorRampPalette(c("darkblue", "dodgerblue", "cyan", "yellow", "red", "darkred"))(100)

# Quick plot check to ensure both loaded correctly
par(mfrow=c(1,2))

# Apply the custom palette using the 'col' argument
plot(mar_raster, main = "GLORYS Annual SST", col = temp_pal)
plot(ter_raster, main = "WorldClim Annual Temp", col = temp_pal)

par(mfrow=c(1,1))
```

3) Overlap Occurrence Records with Correct Ecosystem Rasters
```{r}
# A. Marine Extraction
mar_data <- master_occ %>% filter(Eco == "Mar")
mar_coords <- mar_data %>% dplyr::select(decimalLongitude, decimalLatitude)
mar_data$extracted_temp <- raster::extract(mar_raster, mar_coords)

# B. Terrestrial & Freshwater Extraction
ter_data <- master_occ %>% filter(Eco %in% c("Ter", "Fre"))
ter_coords <- ter_data %>% dplyr::select(decimalLongitude, decimalLatitude)
ter_data$extracted_temp <- raster::extract(ter_raster, ter_coords)

# C. Recombine and clean
sp_occ_combined <- bind_rows(mar_data, ter_data) %>%
  drop_na(extracted_temp) # Drops points that fell completely outside raster boundaries
```

4) Create Per-Species Thermal Summary
```{r}
sp_temp_data <- sp_occ_combined %>% 
  group_by(species, Eco) %>% 
  summarise(
    n_occ = n(),
    min_temp = min(extracted_temp),
    mean_temp = mean(extracted_temp),
    median_temp = median(extracted_temp),
    max_temp = max(extracted_temp),
    mid_point = (max_temp + min_temp) / 2,
    thermal_range = max_temp - min_temp,
    .groups = "drop"
  ) %>% 
  mutate(across(where(is.numeric), round, digits = 3))
```

5) Estimate Temperatures within each H3 Hexagon
We calculate both marine and terrestrial temperatures for every H3 cell, as coastal cells may host both types of species.
```{r}
# Load the unique H3 IDs from harmonized dataset
final_dataset <- read_rds("./stage_4_bind_biotime_bioshifts/final_dataset_lite.rds")
used_h3 <- final_dataset %>% 
  distinct(h3_id) %>%
  mutate(h3_polygon = h3jsr::cell_to_polygon(h3_id))

polygons_sf <- sf::st_as_sf(used_h3$h3_polygon, wkt = used_h3$h3_id) 

# Extract from BOTH rasters (These will now output columns named 'mar_temp' and 'ter_temp')
mar_poly_vals <- raster::extract(mar_raster, polygons_sf, df = TRUE)
ter_poly_vals <- raster::extract(ter_raster, polygons_sf, df = TRUE)

# Summarize Marine H3
mar_h3_summary <- mar_poly_vals %>%
  group_by(ID) %>% 
  summarize(h3_median_mar_temp = round(median(mar_temp, na.rm = TRUE), 3))

# Summarize Terrestrial H3 
ter_h3_summary <- ter_poly_vals %>%
  group_by(ID) %>% 
  summarize(h3_median_ter_temp = round(median(ter_temp, na.rm = TRUE), 3))

# Bind back to H3 IDs
h3_temp_summary <- used_h3 %>%
  mutate(ID = row_number()) %>%
  left_join(mar_h3_summary, by = "ID") %>%
  left_join(ter_h3_summary, by = "ID") %>%
  dplyr::select(h3_id, h3_median_mar_temp, h3_median_ter_temp)
```

6) Calculate Ecosystem-Aware Thermal Position Index
```{r}
# Extract the unique populations from newly built final dataset
pop_info <- final_dataset %>%
  distinct(unique_pop_id, species = species_harmonized, h3_id, Eco)

# Calculate thermal position dynamically
thermal_pos <- pop_info %>%
  left_join(sp_temp_data, by = c("species", "Eco")) %>% 
  left_join(h3_temp_summary, by = "h3_id") %>% 
  filter(n_occ > 1) %>% 
  group_by(unique_pop_id) %>% 
  mutate(
    # dynamically select the correct H3 temperature based on the species' ecosystem
    h3_median_active_temp = ifelse(Eco == "Mar", h3_median_mar_temp, h3_median_ter_temp),
    
    # Calculate the unified thermal position index
    thermal_pos_index = ((h3_median_active_temp - median_temp) / (max_temp - mid_point)) %>% 
                        round(digits = 3)
  ) %>% 
  drop_na(n_occ, thermal_pos_index) %>%
  # Clean up the intermediate calculation columns
  dplyr::select(-h3_median_mar_temp, -h3_median_ter_temp, -h3_median_active_temp)

# Check the spread of the new index
summary(thermal_pos$thermal_pos_index)

# Save the final indices
write_rds(thermal_pos, "./stage_6_thermal_position/thermal_pos_all_ecos.rds")

cat("Calculated thermal positions for", n_distinct(thermal_pos$species), "species.\n")

# Visualize missingness
DataExplorer::plot_missing(thermal_pos)
```

7) Diagnostic Visualizations of Thermal Position
```{r, fig.width=10, fig.height=8}
library(patchwork)

# 1. Recode the Ecosystem abbreviations into full words for the plot
plot_data <- thermal_pos %>%
  mutate(Eco = recode(Eco, 
                      "Mar" = "Marine", 
                      "Ter" = "Terrestrial"))
# 2. Define distinct colors using the NEW full names
eco_colors <- c("Marine" = "dodgerblue", 
                "Terrestrial" = "forestgreen")

# --- PANEL A: Density Distribution ---
# Shows the overall shape and skewness of populations relative to their thermal limits
p_density <- ggplot(plot_data, aes(x = thermal_pos_index, fill = Eco)) +
  # Add reference lines for the Cold Edge (-1), Midpoint (0), and Warm Edge (1)
  geom_vline(xintercept = c(-1, 0, 1), linetype = "dashed", color = "gray50", size = 0.5) +
  geom_density(alpha = 0.6, color = "white", size = 0.5) +
  scale_fill_manual(values = eco_colors) +
  theme_bw(base_size = 14) +
  labs(
    title = "Density distribution of thermal position",
    caption = "Dashed lines represent cold edge (-1), and warm edge (+1)",
    x = "Thermal position",
    y = "Density",
    fill = "Ecosystem"
  ) +
  theme(legend.position = "top")
# Save to disk for records
ggsave("./stage_6_thermal_position/t_pos_eco.png", 
       plot = p_density, width = 12, height = 8, dpi = 300, bg = "white")

cat("Diagnostic plot saved to ./stage_6_thermal_position/diagnostic_thermal_position.png\n")
```
