---
title: "data_overlap_summaries"
author: "Shahar Chaikin"
date: "2025-05-05"
output: html_document
---

Filter datasets according to the ovelapping species list

Libraries
```{r}
library(tidyverse)
```

Load harmonized datasets
```{r}
biotime_harmo=read_rds("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_1_species_harmo\\biotime_harmo.rds")

bioshifts_harmo=read.csv("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_1_species_harmo\\bioshifts_harmo.csv") %>% 
  select(-Gradient)
```

Find species in Bioshifts that are also in BioTIME
```{r}
#First apprroach
#biotime_within_bioshifts_sp_list
biotime_intersect_bioshifts_sp_list=
  intersect(bioshifts_harmo$species_harmonized,biotime_harmo$species_harmonized)

#Second apprroach
biotime_within_bioshifts_sp_list_2=biotime_harmo %>% 
  filter(species_harmonized%in%bioshifts_harmo$species_harmonized)

biotime_within_bioshifts_sp_list_2$species_harmonized %>% n_distinct()

#Yes, both apprroach converge on 4990 species
rm(biotime_within_bioshifts_sp_list_2)
```

Create new subseted datasets according to the new species list
```{r}
#Biotime
biotime_harmo_intersect=biotime_harmo %>% 
  filter(species_harmonized%in%biotime_intersect_bioshifts_sp_list)

##write
#write_rds(biotime_harmo_intersect,"biotime_harmo_intersect.rds")

#Bioshifts
bioshifts_harmo_intersect=bioshifts_harmo %>% 
  filter(species_harmonized%in%biotime_intersect_bioshifts_sp_list)

##write
#write_rds(bioshifts_harmo_intersect,"bioshifts_harmo_intersect.rds")
```

#Read new data and summaries
##Bioshift
```{r}
bioshifts_harmo_intersect=read_rds("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_2_intersect_data/bioshifts_harmo_intersect.rds")

biotime_harmo_intersect=read_rds("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_2_intersect_data/biotime_harmo_intersect.rds") %>% #create unique pop and keep esential columns
  mutate(unique_id=paste(species_harmonized,STUDY_ID,PLOT)) %>% 
  select(unique_id,
         species_harmonized,
         study_id=STUDY_ID,
         plot=PLOT,
         year=YEAR,
         lat=LATITUDE,
         long=LONGITUDE,
         abundance=ABUNDANCE)
```

#BioTIME cleaning for spatial range outliers
```{r}
  biotime_harmo_intersect_spat_cont = biotime_harmo_intersect %>%
  filter(str_count(species_harmonized, "\\S+") >= 2) %>% #remove only genus species
  group_by(unique_id,species_harmonized) %>%  # Group by your unique ID
  # mutate(lat_range = max(lat) - min(lat),  # Calculate latitude range
  #        long_range = max(long) - min(long)) %>%  # Calculate longitude range
  # filter(lat_range <= 1 & long_range <= 1) %>%  # Keep only groups with ranges ≤1°
  mutate(mean_lat_plus_1=mean(lat)+1,
         mean_lat_minus_1=mean(lat)-1,
         mean_long_plus_1=mean(long)+1,
         mean_long_minus_1=mean(long)-1) %>% 
  filter(between(lat,mean_lat_minus_1,mean_lat_plus_1),
         between(long,mean_long_minus_1,mean_long_plus_1)) %>% 
  select(-c(mean_lat_plus_1,mean_lat_minus_1,mean_long_plus_1,mean_long_minus_1)) %>% 
  # Then filter by minimum observations per year
  group_by(unique_id, species_harmonized, year) %>%
  filter(n() >= 3) %>% 
  group_by(unique_id,species_harmonized) %>% 
  filter(sum(abundance)>=1,
         n_distinct(year) >=3) %>% #At least three unique sampled years
  ungroup()
```

#Final intersection with bioshifts
after reducing large latitudinal ranges
```{r}
#filter again for bioshifts
bioshifts_harmo_intersect_spat_cont=bioshifts_harmo_intersect %>% 
  filter(species_harmonized%in%biotime_harmo_intersect_spat_cont$species_harmonized)
```

Write outputs
```{r}
#Biotime_filter
write_rds(biotime_harmo_intersect_spat_cont,"biotime_harmo_intersect_spat_cont.rds")
#Bioshift_filter
write_rds(bioshifts_harmo_intersect_spat_cont,"bioshifts_harmo_intersect_spat_cont.rds")
```

How many fishes?
```{r}
bioshifts_harmo_intersect_spat_cont %>% 
  filter(Class%in%c("Actinopterygii","Chondrichthyes")) %>% 
  distinct(species_harmonized)
```

#Plots
species by ecosystem
```{r}
sp_by_ecosystem=bioshifts_harmo_intersect_spat_cont %>% 
  group_by(Ecosystem) %>% 
  summarise(sp=n_distinct(species_harmonized))

ggplot(sp_by_ecosystem)+
  geom_col(aes(x=Ecosystem,y=sp,fill=Ecosystem))+
  scale_fill_manual(values = c("Marine" = "skyblue", "Terrestrial" = "#6FB74E"))+
  theme_classic()+
  labs(y="Speices")+
  scale_y_continuous(breaks=seq(0,3000,by=200))+
  guides(fill=F)
```

Species by Ecosystem and phyla
```{r}
sp_by_phylum_class=bioshifts_harmo_intersect_spat_cont %>% 
  group_by(Ecosystem,Phylum) %>% 
  summarise(sp=n_distinct(species_harmonized))

ggplot(sp_by_phylum_class)+
  geom_col(aes(x=sp,y=reorder(Phylum,sp),fill=Ecosystem))+
  scale_fill_manual(values = c("Marine" = "skyblue", "Terrestrial" = "#6FB74E"))+
  theme_classic()+
  labs(y="Phylum",x="Species")+
  facet_wrap(~Ecosystem, scales = "fixed") + # Create facets for each Ecosystem
  scale_x_continuous(breaks=seq(0,1000,by=150))+
  guides(fill=F)
```

Species by Ecosystem and Class
```{r}
sp_by_ecosystem_class=bioshifts_harmo_intersect_spat_cont %>% 
  group_by(Ecosystem,Class) %>% 
  summarise(sp=n_distinct(species_harmonized))

ggplot(sp_by_ecosystem_class)+
  geom_col(aes(x=sp,y=reorder(Class,sp),fill=Ecosystem))+
  scale_fill_manual(values = c("Marine" = "skyblue", "Terrestrial" = "#6FB74E"))+
  theme_classic()+
  labs(y="Class",x="Species")+
  facet_wrap(~Ecosystem, scales = "fixed") +
  scale_x_log10()+
  guides(fill=F)
```

Range shift histograms
```{r}
bioshifts_harmo_intersect_spat_cont$Position=factor(bioshifts_harmo_intersect_spat_cont$Position,levels = c("Trailing edge","Centroid","Leading edge"))
ggplot(bioshifts_harmo_intersect_spat_cont)+
  geom_histogram(aes(x=ShiftR,fill=Position))+
  facet_grid(Ecosystem~Position, scales = "fixed")+
  geom_vline(linetype="dashed",xintercept = 0)+
  scale_fill_manual(values = c("Trailing edge" = "#E5474B", "Centroid" = "black","Leading edge"="#92BAE3"))+
  theme_bw()+
  labs(x="Range shift velocity (km/ yr)",y="Count")+
  guides(fill=F)+
  scale_x_continuous(breaks=seq(-50,50,by=20))+
  coord_cartesian(xlim=c(-50,50))
```