---
title: "Limit-Free GBIF Processing Pipeline"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Setup & Credentials

```{r load-libraries-and-credentials}
library(rgbif)
library(CoordinateCleaner)
library(tidyverse)
library(fs)
library(sf)
library(rnaturalearth)
library(GeoThinneR) 
library(arrow) #  For out-of-core processing of massive datasets

# --- ENTER YOUR GBIF CREDENTIALS HERE ---
# GBIF requires credentials for unlimited downloads.
Sys.setenv(
  GBIF_USER = "shahar_chaikin",
  GBIF_PWD  = "o08*ZE!7m1Cp",
  GBIF_EMAIL = "shahar710@gmail.com"
)
```

2. Prepare Species List & Fetch Data
Instead of relying on literal string matches which can fail in GBIF, we use the rgbif backbone matching tool to get the exact taxonKey for each of your ~730 species.

```{r}
# Load the target species list generated in Stage 4
target_species <- read.csv("./stage_4_bind_biotime_bioshifts/target_species_for_occurrences.csv")
```

```{r}
# 1. Extract and clean the existing GBIF keys
taxa_keys <- target_species %>%
  filter(!is.na(taxonID)) %>%
  mutate(usageKey = as.numeric(str_remove(taxonID, "GBIF:"))) %>%
  distinct(usageKey) %>%
  pull(usageKey)

cat("Successfully extracted", length(taxa_keys), "unique GBIF keys.\n")

# 2. Request the massive bulk download
gbif_download <- occ_download(
  pred_in("taxonKey", taxa_keys),
  pred("hasCoordinate", TRUE),
  pred_gte("year", 1900), 
  format = "SIMPLE_CSV"
)

# 3. Wait for the server to finish generating the file
occ_download_wait(gbif_download)

# 4. DOWNLOAD ONLY - DO NOT IMPORT
# This saves the massive .zip file to your working directory.
occ_download_get(gbif_download, path = "D:/Shahar_Chaikin_phd/sRED/occ_05_2026/gbif_raw_data")

cat("Download complete! STOP HERE. Go to the './gbif_raw_data' folder on your computer and manually unzip the file before proceeding to Step 3.\n")
```

3. Arrow Partitioning (The 168GB Rescue)
Once we have manually unzipped the file (which will yield a massive .csv), we use arrow to scan it on the hard drive and slice it into hundreds of tiny files, grouped by species.
```{r}
# 3. Arrow Partitioning (The 168GB Rescue)
library(purrr)

raw_csv_path <- "D:/Shahar_Chaikin_phd/sRED/occ_05_2026/gbif_raw_data/0002281-260519110011954.csv"

# 1. Instantly extract the column headers
header <- strsplit(readLines(raw_csv_path, n = 1), "\t")[[1]]
header <- gsub('^"|"$', '', header) 

# 2. Build the Bulletproof Schema (Everything is a string)
string_types <- map(setNames(header, header), ~utf8())
bulletproof_schema <- do.call(schema, string_types)

# 3. Open the dataset (Using ONLY native Arrow options)
massive_csv <- open_dataset(
  raw_csv_path, 
  format = "csv", 
  delimiter = "\t",
  schema = bulletproof_schema,
  skip_rows = 1,
  null_values = c("", "NA", "null", " "),
  quoting = FALSE, 
  newlines_in_values = TRUE 
)

massive_csv %>%
  select(
    speciesKey, 
    decimalLongitude, 
    decimalLatitude, 
    occurrenceStatus, 
    basisOfRecord, 
    coordinateUncertaintyInMeters
  ) %>%
  filter(
    !is.na(decimalLongitude) & decimalLongitude != "",
    !is.na(decimalLatitude) & decimalLatitude != "",
    !is.na(speciesKey) & speciesKey != ""
  ) %>%
  mutate(
    speciesKey = as.integer(speciesKey),
    decimalLongitude = as.numeric(decimalLongitude),
    decimalLatitude = as.numeric(decimalLatitude)
  ) %>%
  
  # THE FIX: We completely removed group_by(speciesKey)!
  # Arrow will now write sequentially at maximum hard drive speed.
  write_dataset(path = "D:/Shahar_Chaikin_phd/sRED/occ_05_2026/gbif_flat_parquet", format = "parquet")

cat("Linear Parquet database created!\n")
```

4. Define the Spatial Mask & Processing Function
Here we load a high-quality global land shapefile and define the processing function that will filter, mask, thin, and save each species to its own .rds file.

```{r}
# Load a global physical land polygon for masking
land_sf <- ne_download(scale = 50, type = "land", category = "physical", returnclass = "sf") %>%
  st_make_valid() %>%
  st_union()

process_and_save <- function(df, target_species, target_eco) {
  
  output_dir <- "./stage_5_coords_ext/sp_occ_cleaned"
  dir_create(output_dir)
  
  safe_name <- str_replace_all(target_species, " ", "_")
  output_file <- path(output_dir, paste0(safe_name, ".rds"))
  
  message(sprintf("Processing: %s [%s]", target_species, target_eco))
  
  # A. Tidyverse Metadata Filters
  occ_filtered <- df %>%
    mutate(species = target_species) %>%
    mutate(coordinateUncertaintyInMeters = suppressWarnings(as.numeric(coordinateUncertaintyInMeters))) %>%
    mutate(
      occurrenceStatus = if(!"occurrenceStatus" %in% names(.)) "PRESENT" else occurrenceStatus,
      coordinateUncertaintyInMeters = if(!"coordinateUncertaintyInMeters" %in% names(.)) NA else coordinateUncertaintyInMeters,
      basisOfRecord = if(!"basisOfRecord" %in% names(.)) "UNKNOWN" else basisOfRecord
    ) %>%
    filter(
      str_to_upper(occurrenceStatus) == "PRESENT",
      str_to_upper(basisOfRecord) != "FOSSIL_SPECIMEN",
      is.na(coordinateUncertaintyInMeters) | coordinateUncertaintyInMeters <= 50000
    ) %>%
    distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE)
  
  if(nrow(occ_filtered) == 0) return(invisible(FALSE))
  
  # B. CoordinateCleaner 
  occ_clean <- clean_coordinates(
    x = occ_filtered,
    lon = "decimalLongitude", lat = "decimalLatitude",
    tests = c("zeros", "capitals", "centroids", "institutions"), 
    value = "clean", verbose = FALSE
  )
  
  rm(occ_filtered)
  gc()
  
  if(nrow(occ_clean) == 0) return(invisible(FALSE))
  
  # C. Strict sf Land/Sea Masking (THE FIX: Bulletproof Global Intersect)
  occ_sf <- st_as_sf(occ_clean, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE)
  
  sf_use_s2(FALSE)
  
  # We removed st_crop. We just ask the heavy global polygon directly.
  # Wrapped in a tryCatch to automatically handle any rogue geometry errors.
  on_land <- tryCatch({
    lengths(st_intersects(occ_sf, land_sf)) > 0
  }, error = function(e) {
    # If flat math fails, turn spherical math (S2) back on for a safe fallback!
    sf_use_s2(TRUE)
    lengths(st_intersects(occ_sf, sf::st_make_valid(land_sf))) > 0
  })
  
  sf_use_s2(TRUE)
  
  if (target_eco == "Mar") {
    occ_clean <- occ_clean[!on_land, ]
  } else if (target_eco %in% c("Ter", "Fre")) {
    occ_clean <- occ_clean[on_land, ]
  }
  
  rm(occ_sf, on_land)
  gc()
  
  if(nrow(occ_clean) == 0) return(invisible(FALSE))
  
  # D. Spatial Thinning 
  occ_thinned <- tryCatch({
    thinned_obj <- GeoThinneR::thin_points(
      data = occ_clean,
      lon_col = "decimalLongitude", lat_col = "decimalLatitude",
      method = "grid", 
      resolution = 0.1, 
      trials = 1
    )
    occ_clean[thinned_obj$retained[[1]], ]
  }, error = function(e) {
    message(sprintf("  -> Thinning Error: %s", e$message))
    return(NULL)
  })
  
  if (is.null(occ_thinned) || nrow(occ_thinned) == 0) return(invisible(FALSE))
  
  # E. Save to file 
  write_rds(occ_thinned, output_file)
  message(sprintf("  -> SUCCESS: Saved %s clean records.", nrow(occ_thinned)))
  
  return(invisible(TRUE))
}
```

5. Execute Pipeline on Parquet Partitions
Finally, we loop through the tiny Parquet folders created by arrow, pull the correct species metadata from your target_species list, and clean the data one by one.
```{r}
library(furrr)
library(arrow)

species_lookup <- target_species %>%
  filter(!is.na(taxonID)) %>%
  mutate(speciesKey = as.integer(str_remove(taxonID, "GBIF:"))) %>%
  select(speciesKey, species_harmonized, Eco)

# THE FIX: Hardcode this to exactly 2 workers to maximize RAM per core
plan(multisession, workers = 2)

future_walk(species_lookup$speciesKey, function(current_key) {
  
  fast_parquet <- arrow::open_dataset("./stage_5_coords_ext/fast_parquet", format = "parquet")
  
  meta <- species_lookup %>% filter(speciesKey == current_key)
  
  if(nrow(meta) > 0) {
    target_species_name <- meta$species_harmonized[1]
    safe_name <- str_replace_all(target_species_name, " ", "_")
    expected_output <- path("./stage_5_coords_ext/sp_occ_cleaned", paste0(safe_name, ".rds"))
    
    if (file_exists(expected_output)) {
      return() 
    }
    
    current_df <- fast_parquet %>%
      filter(speciesKey == current_key) %>%
      collect()
    
    if(nrow(current_df) > 0) {
      process_and_save(current_df, target_species_name, meta$Eco[1])
    }
    
    # Final worker trash-empty
    gc()
  }
  
}, .options = furrr_options(
  seed = TRUE, 
  packages = c("dplyr", "sf", "CoordinateCleaner", "stringr", "fs", "readr")
), .progress = TRUE)

cat("Parallel Processing complete.\n")
```

# 6. Gather and Bind Cleaned Occurrences
```{r}
library(purrr)

# 1. Point to the directory where process_and_save() dropped the files
cleaned_dir <- "./stage_5_coords_ext/sp_occ_cleaned"

# 2. Get a list of all the individual .rds files
cleaned_files <- dir_ls(cleaned_dir, glob = "*.rds")

# 3. Read all files and bind them into a single massive dataframe
# map_dfr is extremely fast and automatically stacks the rows
master_occ_data <- cleaned_files %>%
  map_dfr(read_rds)

# 4. Save the master dataset for the Thermal Position script
# Saving as .rds is highly recommended over .csv because it preserves spatial/numeric column types
write_rds(master_occ_data, "./stage_5_coords_ext/master_cleaned_occurrences.rds")

cat("Successfully gathered", nrow(master_occ_data), "total occurrences across", 
    n_distinct(master_occ_data$species), "species.\n")
```

