---
title: "download_occ_simple"
author: "Shahar Chaikin"
date: "2024-12-17"
output: html_document
---
As I already downloaded occurrences using the abundance dataset, here I will only focus on species that are unique to the biomass dataset.

Libraries
```{r}
library(tidyverse)
library(rgbif)
library(CoordinateCleaner)
```

What is the biomass species list?
```{r}
sp_list = read.csv("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_7_occ_extraction\\sp_list.csv") %>%
  mutate(usageKey = as.numeric(str_extract(usageKey, "\\d+"))) %>%
  distinct(usageKey,species_harmonized)
```

GBIF credentials
```{r}
user="shahar_chaikin"
pwd="o08*ZE!7m1Cp"
email="shahar710@gmail.com"
```

Occurrence extraction function
```{r}
# Function to download GBIF data for a vector of usageKeys and return a combined dataframe
download_gbif_data <- function(usageKeys, user, pwd, email, format = "SIMPLE_CSV") {
  
  # Create an empty data frame to store results
  combined_data <- data.frame()
  
  for (i in 1:length(usageKeys)) {
    
    # Build the query with the current usageKey
    pred_taxon <- pred("taxonKey", usageKeys[i])
    
    # Download data with wait and error handling
    download <- tryCatch({
      occ_download(pred_taxon,
                  pred("hasGeospatialIssue", FALSE),
                  pred("hasCoordinate", TRUE),
                  pred("occurrenceStatus", "PRESENT"),
                  pred_gte("year", 1900),
                  #pred_lte("coordinateUncertaintyInMeters", 10000),
                  user = user,
                  pwd = pwd,
                  email = email,
                  format = format)
    }, error = function(e) {
      warning(paste("Error downloading data for usageKey:", usageKeys[i]))
      return(NULL)
    })
    
    # Check if download was successful
    if (!is.null(download)) {
      # Wait for download to finish
      occ_download_wait(download)
      # Get and import data
      current_data <- occ_download_get(download) |> occ_download_import()
      # Bind the current data to the combined data frame
      combined_data <- rbind(combined_data, current_data) 
    }
  }
  
  return(combined_data)
}
```


#1) Download
Download all occ in the wanted destination
```{r}
setwd("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\stage_7_occ_extraction\\occ_raw/")
```

##1.1) a bulk
Done
```{r}
a_bulk=download_gbif_data(usageKeys =as.character(sp_list$usageKey[1:2]),
                   user = user,
                   pwd = pwd,
                   email = email)
write_rds(a_bulk,"a_bulk_shahar.rds")
```

Read file
```{r}
raw_occ_no_unc_limit=read_rds("C:\\Users\\User\\Desktop\\research\\working groups\\iDiv_2024_25_sRED\\R\\sRED_demog_range_shifts\\")
```

#2) Remove occ. on land
```{r}
library(sf)

# Load your data
data <- raw_occ_no_unc_limit 

# Create an sf object from your data
occurrences_sf <- sf::st_as_sf(data, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) 

# Load the land shapefile
land_sf <- sf::st_read("C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\ne_10m_land\\ne_10m_land.shp") 

# Check for occurrences on land
occurrences_on_land <- lengths(sf::st_intersects(occurrences_sf, land_sf)) > 0 

# Flag occurrences on land
data$on_land <- occurrences_on_land

# Remove occurrences on land
data_filtered <- data %>% 
  filter(!on_land)

#How many were removed?
length(data$on_land)-length(data_filtered$on_land)

# Output the filtered data
#write_rds(data_filtered, "occ_filtered_sea_no_unc_limit_biomass_public_MEDITS_add.rds") 
occ_filtered_sea_no_unc_limit=data_filtered
rm(data_filtered)
rm(data)
rm(raw_occ_no_unc_limit)
rm(occurrences_sf)
```

Read up till here
```{r}
occ_filtered_sea_no_unc_limit=read_rds("C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\gbif_bulks_no_uncerteinty_limit_biomass/occ_filtered_sea_no_unc_limit_biomass.rds")

#Explore the 'coordinateUncertaintyInMeters' variable
options(scipen = 999)
DataExplorer::plot_missing(occ_filtered_sea_no_unc_limit)

discard(occ_filtered_sea_no_unc_limit$coordinateUncertaintyInMeters,is.na) %>% 
  range
```
#3) additional filters
```{r}
occ_filtered_fine_no_unc_limit=occ_filtered_sea_no_unc_limit %>% 
  filter(
    #remove fossils and living specimens from zoos
    !basisOfRecord %in% c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"),
    #GeoLocate centroids or other default. remove them because usually the            uncertainty is larger than stated
    !coordinateUncertaintyInMeters %in% c(301, 3036, 999, 9999),
    #When uncertainty in known remove occurrences with accuracy bigger then 50 km
    #A balance between quality and quantity when possible
    is.na(coordinateUncertaintyInMeters) | 
           coordinateUncertaintyInMeters <= 50000,
    !decimalLatitude == 0 | !decimalLongitude == 0) %>%
    #drop_na(coordinateUncertaintyInMeters) %>% 
   #Remove coords in the center of countries
   #Remove coords in capitals
   #Remove coords near GBIF HQ
   #Remove coords near zoological institutions
    cc_cen(buffer = 2000, lon = "decimalLongitude", lat = "decimalLatitude") %>%
    cc_cap(buffer = 2000, lon = "decimalLongitude", lat = "decimalLatitude") %>%
    cc_inst(buffer = 2000, lon = "decimalLongitude", lat = "decimalLatitude") %>%
    cc_gbif(buffer = 2000, lon = "decimalLongitude", lat = "decimalLatitude") %>%
    #Remove duplicates
    distinct(decimalLongitude,
             decimalLatitude,
             speciesKey,
             datasetKey,
             #Keep all other columns
             .keep_all = TRUE)

write_rds(occ_filtered_fine_no_unc_limit,"occ_filtered_fine_no_unc_limit.rds")
```

Read up till here if needed
```{r eval=FALSE, include=FALSE}
occ_filtered_fine_no_unc_limit=read_rds("C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\gbif_bulks_no_uncerteinty_limit_biomass/occ_filtered_sea_no_unc_limit_biomass_public_MEDITS_add.rds")
```

Return species names according to the harmonization file and the usageKey
```{r}
#How many distinct species in the filtered data?
occ_filtered_fine_no_unc_limit$speciesKey %>% n_distinct()

#Which species are in the harmonized file but not in the occ file?
setdiff(unique(sp_list$usageKey),unique(occ_filtered_fine_no_unc_limit$speciesKey))
#Almost half of the species downloaded now don't have a uasagekey that matches the list provided.

#Return the harmonized species names according to the useageKey
sp_list_cleaned_occ = occ_filtered_fine_no_unc_limit %>%
  select(usageKey = speciesKey,
         decimalLatitude,
         decimalLongitude) %>%
  left_join(sp_list %>% 
              select(usageKey,
                     species = scientificName) %>% 
              #Handle duplicate rows such as in Aetomylaeus bovinus
              distinct(usageKey,species),
            by = "usageKey")
#write_rds(sp_list_cleaned_occ,"sp_occ_lite_no_uncertainty_limits_biomass_addition_public_MEDITS_add.rds")

#How many species have occurrences?
sp_list$scientificName %>% n_distinct()
sp_list_cleaned_occ$usageKey %>% n_distinct()
# Is theremore than one usageKey for some species
sp_list_cleaned_occ %>% 
  group_by(species) %>% 
  summarise(n_usageKey=n_distinct(usageKey)) %>% 
  filter(n_usageKey>1)
#no

#How many occurrences per species?
occ_per_sp_before_spthin=sp_list_cleaned_occ %>% 
  group_by(species) %>% 
  summarise(n_occ=n_distinct(decimalLatitude,
                             decimalLongitude))
#summary sts.
#Plot
occ_per_sp_before_spthin$n_occ %>% quantile() 
ggplot(occ_per_sp_before_spthin)+
  geom_violin(aes(y=n_occ,x=""),
              fill="grey")+
  geom_jitter(aes(y=n_occ,x=""),
              width=0.1,
              alpha=0.3)+
  geom_boxplot(aes(y=n_occ,x=""),
               width=0.4,
               size=1,
               fill="green",
               alpha=0.7)+
  scale_y_log10()+
  labs(y="Occurrences",x="")+
  theme_bw()
```

#4) Spatial thinning
To be looped at the species level
```{r eval=FALSE, include=FALSE}
library(spThin)
rm(occ_filtered_fine_no_unc_limit)
for (i in occ_per_sp_before_spthin %>% 
     drop_na(species) %>% 
     select(species) %>% 
     pull() %>% 
     unique()){
 spThin::thin(
  loc.data = sp_list_cleaned_occ %>% 
    filter(species%in%i),
  lat.col = "decimalLatitude",
  long.col = "decimalLongitude",
  spec.col = "species",
  #min NND of 10km (try something else to see if it runs )
  thin.par = 10,
  #One repetition per species
  reps = 1,
  #One dataset per species is enough - the df with most occ. is written
  max.files = 1,
  out.dir = "C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\spthin/public_MEDITS_addition/",
  out.base = paste(i))
}
```

Load all thin spp. files
```{r}
# Get a list of all .rds files in the folder
thin_file_paths <- list.files(path = "C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\spthin\\public_MEDITS_addition/", pattern = "*.csv", full.names = TRUE)

# Read all .rds files into a list of data frames
thin_data_list <- map(thin_file_paths, read.csv)

# Bind all rows of the data frames in the list
thin_sp_occ <- bind_rows(thin_data_list)
#write_rds(thin_sp_occ,"thin_sp_occ_no_uncertainty_limits_biomass_addition_public_MEDITS_add.rds")
```

Read result if needed
```{r}
thin_sp_occ=read_rds("C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\stage_2_spat_therm_position\\spthin\\with_no_uncerteinty_limit_biomass\\thin_sp_occ_no_uncertainty_limits_biomass_addition.rds")
#How many species?
thin_sp_occ$species %>% n_distinct()
```

Investigate the final occurrence dataset
```{r}
#How many species?
thin_sp_all_sp$species %>% n_distinct()
#Occurrences per species
thin_sp_occ_summary=thin_sp_all_sp %>% 
  group_by(species) %>% 
  summarise(n_occ=n_distinct(decimalLongitude,
                            decimalLatitude))

#Percentiles of occurrences
thin_sp_occ_summary$n_occ %>% quantile()
#Plot: Occurrences per species
ggplot(thin_sp_all_sp%>% 
  group_by(species) %>% 
  summarise(n_occ=n_distinct(decimalLatitude,
                             decimalLongitude)))+
  geom_violin(aes(x=n_occ,y=""),
              fill="grey",
              alpha=0.8,
              size=NA)+
  geom_jitter(aes(x=n_occ,y=""),
              width=0.1,
              alpha=0.5,
              color="grey")+
  geom_boxplot(aes(x=n_occ,y=""),
               width=0.4,
               size=0.8,
               fill="green",
               alpha=0.4,
               notch=T,
               outliers = F)+
  scale_x_log10()+
  labs(x="Occurrences",y="Species",title = paste("Occurrence per species (Spp. = ",
                                                 thin_sp_all_sp$species %>% 
                                                   n_distinct(),
                                                 ")"))+
  theme_bw()
```

